---
title: 'Pushed expansions and sources of variation- individual-based model, data analysis'
author: "Maxime Dahirel & Chloé Guicharnaud"
date:
output:
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

```{r load-packages}
library(ordinal)
library(arm)
library(matrixStats)
library(MuMIn)
library(tidyverse)
library(cowplot)
library(patchwork)

library(ggtext)


library(mgcv)
library(gratia)

library(here)
```

# Introduction

```{r}
raw_data <- read_csv(here("NetLogo_output","simulation_output.csv"))
```

We rewrite variance variables to be in 0/1 mode to not be distracted

```{r}
data<- raw_data  %>% 
  mutate(is.VP_dmax=as.numeric(VP_logit_dmax >0),
           is.VP_DDD=as.numeric(VP_slope >0))
```


# making sub data tables

a table with summary variables at the edge (we define edge as patch <5 patchs from tip, see suppl for justification)
```{r}
tab_edge<-data %>% 
  group_by(replicateID,ticks) %>% 
  mutate(front=max(pxcor*(N_postdispersal>0))) %>% 
  filter(pxcor > front-5 ) %>%
  ungroup() %>% 
  group_by(replicateID,ticks,dispersal_mortality,heritability,is.VP_dmax,is.VP_DDD) %>% 
  summarise(
    mean_dmax=weighted.mean(mean_dmax,N_predispersal,na.rm=TRUE),
    mean_davg_0_K=weighted.mean(mean_davg_0_K,N_predispersal,na.rm=TRUE),
    mean_d0=weighted.mean(mean_d0,N_predispersal,na.rm=TRUE),
    mean_avgslope=weighted.mean(mean_slopeA_0_avg,N_predispersal,na.rm=TRUE),
    mean_Kslope=weighted.mean(mean_slopeA_0_K,N_predispersal,na.rm=TRUE),
    mean_uncond=weighted.mean(mean_uncond_0_K,N_predispersal,na.rm=TRUE),
    front=mean(front)
  ) %>% 
  ungroup() %>% 
  group_by(replicateID) %>% 
  ungroup() %>% 
  pivot_wider(id_cols = c(replicateID,dispersal_mortality,heritability,is.VP_dmax,is.VP_DDD),
              names_from=ticks,
              values_from=c(mean_dmax, mean_davg_0_K, mean_d0, 
                            mean_avgslope,mean_Kslope,mean_uncond,
                            front))
```


we also extract separately variables at t0
```{r}
tab_start<-data %>% select(
  start_dmax,
  start_davg_0_K,
  start_d0,
  start_avgslope=start_slopeA_0_avg,
  start_Kslope=start_slopeA_0_K,
  start_uncond=start_uncond_0_K,
  replicateID
) %>% 
  distinct()

```

finally we run another extraction to get information on genetic diversity decay, time to fixation
(earliest observation of front (defined as above) only containing one of the two neutral alleles
)
```{r}
tab_genet <- data %>% 
  group_by(replicateID,ticks) %>% 
  mutate(front=max(pxcor*(N_postdispersal>0))) %>% 
  filter(pxcor > front-5 ) %>%
  ungroup() %>% 
  group_by(replicateID,ticks,dispersal_mortality,heritability,is.VP_dmax,is.VP_DDD) %>% 
  summarise(
    N_allele0=sum(N_allele0_post),
    N_allele1=sum(N_allele1_post)
  ) %>% 
  ungroup() %>% 
  mutate(edge_fix= N_allele0==0 | N_allele1==0) %>% 
  group_by(replicateID) %>% 
  mutate(time_to_fix=case_when(subset(edge_fix,ticks==20)==1 ~ 20,
                               subset(edge_fix,ticks==40)==1 ~ 40,
                               subset(edge_fix,ticks==60)==1 ~ 60,
                               subset(edge_fix,ticks==80)==1 ~ 80,
                               subset(edge_fix,ticks==100)==1 ~ 100,
                               subset(edge_fix,ticks==120)==1 ~ 120,
                               T~121)##not fixed yet at end of experiment
         ) %>% 
  ungroup() %>% 
  group_by(replicateID,dispersal_mortality,heritability,is.VP_dmax,is.VP_DDD) %>% 
  summarise(time_to_fix=unique(time_to_fix)) %>% 
  ungroup() %>% 
  mutate(fixinterval=ordered(time_to_fix,
                             levels=sort(unique(time_to_fix))
                             )
         )
```

now we can mix and combine these datasets for analyses

# analysis of speed and genetic dynamics


```{r}
tab <-tab_edge %>% 
  left_join(tab_start) %>% 
  left_join(tab_genet) %>% 
  mutate(mortality=factor(dispersal_mortality)) %>% 
  mutate(advance=front_120-front_100) 
```

for both velocity and genetics, we compare xx models:

- with and without the density-dependency variable (is knowledge of DDD needed to predict?)
- at g100 and at the start of the experiment (is knowledge of evolution needed?)


# Figure 1: overview of the dispersal function

```{r fig 1}

p1_tab <- tibble(x=c(0:120)/100) %>% 
  expand_grid(dmax=c(0.2,0.5,0.7),
              a=c(-10,0,10),
              b=c(0.2,0.5,1.1)) %>% 
  mutate(group=paste(dmax,a,b)) %>% 
  mutate(d=dmax/(1+exp(-a*(x-b))))

p1_1_dmax <- p1_tab %>% 
  filter(b==0.5 & a==10)%>% 
  ggplot()+
  geom_line(aes(x,d,group=group))+
  geom_textbox(data=tibble(dmax=unique(p1_tab$dmax)),
                 aes(x=1,
                 y=dmax-0.05,
                 label=paste("*d<sub>max</sub>* = ",dmax)),
               width = grid::unit(0.3, "npc")
                 )+
  scale_x_continuous("N/K")+
  scale_y_continuous("Dispersal probability")+
  theme_bw()

p1_2_alpha <- p1_tab %>% 
  filter(b==0.5 & dmax==0.7)%>% 
  ggplot()+
  geom_line(aes(x,d,group=group))+
  geom_textbox(data=tibble(b=0.5,
                           dmax=0.7,
                           a=unique(p1_tab$a)),
               aes(x=1,
                 y=dmax/(1+exp(-a*(1-b))),
                 label=paste("*&alpha;* =",a)),
               width = grid::unit(0.22, "npc"))+
  scale_x_continuous("N/K")+
  scale_y_continuous("")+
  theme_bw()

p1_3_beta <- p1_tab %>% 
  filter(a==10 & dmax==0.7)%>% 
  ggplot()+
  geom_line(aes(x,d,group=group))+
  geom_textbox(data=tibble(b=unique(p1_tab$b),
                           dmax=0.7,
                           a=10),
               aes(x=b,
                 y=0.4,
                 label=paste("*&beta;* =",b)),
               width = grid::unit(0.15, "npc"))+
  scale_x_continuous("N/K")+
  scale_y_continuous("")+
  theme_bw()


p1_tab2 <- tibble(dmax=c(0.734,0.8),
              a=c(20,6),
              b=c(0.2,0.6)) %>%
  expand_grid(x=rep(c(0:10)/10,2)) %>% 
  mutate(d=dmax/(1+exp(-a*(x-b))))


p1_4<- p1_tab2 %>% 
  ggplot(aes(x=x,y=d))+
  geom_line(aes(col=factor(dmax)))+
  geom_point(aes(col=factor(dmax),shape=factor(dmax)),size=5,fill="white")+
  scale_x_continuous("N/K")+
  scale_y_continuous("Dispersal probability")+
  scale_shape_discrete()+
  theme_bw()

p1_5 <- p1_tab2 %>% 
  group_by(dmax) %>% 
  summarise(d0=subset(d,x==0),
            dK=subset(d,x==1),
            davg=mean(d)) %>% 
  pivot_longer(cols=c(d0,dK,davg)) %>% 
  mutate(name=fct_recode(name,
                         `*d<sub>0</sub>*`="d0",
                         `*d<sub>avg</sub>*`="davg",
                         `*d<sub>K</sub>*`="dK")) %>% 
  ggplot()+
  geom_point(aes(name,value,col=factor(dmax),shape=factor(dmax)),
             size=5,fill="white",position=position_dodge(width=0.4))+
  scale_x_discrete("dispersal variable")+
  scale_y_continuous("value")+
  scale_shape_discrete()+
  theme_bw()+
  theme(axis.text.x = element_markdown())

(p1_1_dmax|p1_2_alpha|p1_3_beta)/(p1_4+p1_5)&
  theme(legend.position = "none") & 
  plot_annotation(tag_levels = "A")

```


# Do the models for velocity and neutral genetic diversity dynamics

## velocity

### models

```{r}
mod_100=gam(cbind(advance,20-advance)~
              s(mean_d0_100,by=mortality)+
              ##read up on smooths before final model to be sure to use them correctly
              s(mean_avgslope_100,by=mortality)+
              mortality, 
            family=binomial,data=tab, method = "REML")
###does better than the nonlinear at d0=nearly 0 because d0 is an average
### even when d0=0, some individuals may not be at d0

mod_100_K=gam(cbind(advance,20-advance)~
              s(mean_d0_100,by=mortality)+
              s(mean_Kslope_100,by=mortality)+
              mortality, 
            family=binomial,data=tab, method = "REML")

mod_100_noDDD=gam(cbind(advance,20-advance)~
              s(mean_d0_100,by=mortality)+
              mortality, 
            family=binomial,data=tab, method = "REML")


mod_00=gam(cbind(advance,20-advance)~
              s(start_d0,by=mortality)+
              s(start_avgslope,by=mortality)+
              mortality, 
            family=binomial,data=tab, method = "REML")

mod_00_K=gam(cbind(advance,20-advance)~
              s(start_d0,by=mortality)+
              s(start_Kslope,by=mortality)+
              mortality, 
            family=binomial,data=tab, method = "REML")

mod_00_noDDD=gam(cbind(advance,20-advance)~
              s(start_d0,by=mortality)+
              mortality, 
            family=binomial,data=tab, method = "REML")

mod_null=gam(cbind(advance,20-advance)~
              mortality, 
            family=binomial,data=tab, method = "REML")
```

seems like data are underdispersed relative to the binomial, for all four models

```{r}
## function from ben bolker FAQ https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#overdispersion
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

overdisp_fun(mod_100) ##the p value is for a test of overdispersion, not under look at the ratio
```



see comment from bob carpenter here: https://discourse.mc-stan.org/t/underdispersed-binomial-glm/1540
the idea that if underdispersed relative to binomial, then not the result of independent bernoulli trials
which make sense, since proba of advancing to N+1 depends on # of individuals that advanced to N
qualitative ccl are the same anyway though
an ordinal model may be appropriate instead
see eg https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#underdispersion
so let's go with that as a check (supplement?)

```{r}

#library(cmdstanr)
#library(brms)
#
#tab_brm <- tab %>%
#  mutate(scale_d0_100=scale(mean_d0_100)[,1], #standardisation only used for the brms check
#         scale_d0_start=scale(start_d0)[,1],
#         s_avgslope_100=mean_avgslope_100/sd(mean_avgslope_100),
#         s_avgslope_start=start_avgslope/sd(start_avgslope),
#         s_Kslope_100=mean_Kslope_100/sd(mean_Kslope_100),
#         s_Kslope_start=start_Kslope/sd(start_Kslope))
#
#prior<-c(
#  set_prior("normal(0,1.5)",class="b")
#)
#
#mod_100_ordinal=brm(bf((advance+1)~ #because the first category must be 1
#          (scale_d0_100+s_avgslope_100)*mortality),
#        family=sratio,data=tab_brm,prior=prior,
#        chains=4,seed=42,iter=2000,backend="cmdstanr")

## easy to expand to the otehr models
```

### model selection

```{r}
gam_dev.expl=function(mod){
            return(summary(mod)$dev.expl)
}

gam_r2=function(mod){
            return(summary(mod)$r.sq)
          }


modsel_velocity <-model.sel(list(mod_100=mod_100,
               mod_100_K=mod_100_K,
               mod_100_noDDD=mod_100_noDDD,
               mod_00=mod_00,
               mod_00_K=mod_00_K,
               mod_00_noDDD=mod_00_noDDD,
               mod_null=mod_null),extra=list(gam_dev.expl,gam_r2)) %>%
  rownames_to_column(var="model_name") %>% 
  as_tibble() %>% 
             mutate(is.evo=case_when(str_detect(model_name,"_100")~"after evolution (t = 100)",
                                     T~"before evolution (t = 0)"),
                    is.DDD=case_when(str_detect(model_name,"_noDDD")~"ignored",
                                     str_detect(model_name,"_K")~"*&Delta;<sub>K-0</sub>* (shape-indep.)",
                                     T~"*&Delta;<sub>avg-0</sub>* (shape-dep.)"))  %>%
  mutate(is.evo=fct_relevel(is.evo,"before evolution (t = 0)",after=0)) %>%
  mutate(is.DDD=fct_relevel(is.DDD,"ignored",after=0))        ##zero doubt the evol model is better
```

## diversity

### models

```{r}
###do double checks for logit vs cloglog and adjust text in manuscript
mod_div_00=clm(fixinterval~
                     mortality*(start_d0+start_avgslope),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance)    


mod_div_00=clm(fixinterval~
                     mortality+mortality:(start_d0+start_avgslope),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance)    

mod_div_00_K=clm(fixinterval~
                     mortality+mortality:(start_d0+start_Kslope),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance) 

mod_div_00_noDDD=clm(fixinterval~
                     mortality+mortality:(start_d0),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance) 

mod_div_100=clm(fixinterval~
                     mortality+mortality:(mean_d0_100+mean_avgslope_100),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance)    

mod_div_100_K=clm(fixinterval~
                     mortality+mortality:(mean_d0_100+mean_Kslope_100),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance) 

mod_div_100_noDDD=clm(fixinterval~
                     mortality+mortality:(mean_d0_100),
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance) 

mod_div_null=clm(fixinterval~
                     mortality,
        data=filter(tab,front_120>0),
        link="cloglog")#one front did not advance) 


## the cloglog/proportional hazards more "traditional" for survival data
## also easy to change family and see it is a better fit
```



### model selection table

```{r}
clm_pseudo_r2<- function(model,data=tab){
  pp<-predict(model,newdata=data %>% select(-fixinterval))$fit
  preds<-1*pp[,1]+2*pp[,2]+3*pp[,3]+4*pp[,4]+pp[,5]+6*pp[,6]+7*pp[,7]
  obs<-as.numeric(tab$fixinterval)
  return(summary(lm(obs~preds))$r.squared)
}

modsel_diversity <- model.sel(list(mod_div_100=mod_div_100,
               mod_div_100_K=mod_div_100_K,
               mod_div_100_noDDD=mod_div_100_noDDD,
               mod_div_00=mod_div_00,
               mod_div_00_K=mod_div_00_K,
               mod_div_00_noDDD=mod_div_00_noDDD),extra=list(clm_pseudo_r2)) %>%
  rownames_to_column(var="model_name") %>% 
  as_tibble() %>% 
             mutate(is.evo=case_when(str_detect(model_name,"_100")~"after evolution (t = 100)",
                                     T~"before evolution (t = 0)"),
                    is.DDD=case_when(str_detect(model_name,"_noDDD")~"ignored",
                                     str_detect(model_name,"_K")~"*&Delta;<sub>K-0</sub>* (shape-indep.)",
                                     T~"*&Delta;<sub>avg-0</sub>* (shape-dep.)")) %>%
  mutate(is.evo=fct_relevel(is.evo,"before evolution (t = 0)",after=0))%>%
  mutate(is.DDD=fct_relevel(is.DDD,"ignored",after=0))    

```

# figure 2 : model selection

```{r}

p2_1 <- modsel_velocity %>%
  filter(model_name!="mod_null") %>%
  ggplot()+
  geom_point(aes(is.evo,gam_r2,
                         col=is.DDD),
                     size=3)+
  geom_line(aes(is.evo,gam_r2,
                group=is.DDD,
                col=is.DDD))+
  geom_hline(yintercept=summary(mod_null)$r.sq,lty=2)+ ## the aicc of the mortality only model
  scale_y_continuous("R^2")+
  labs(title="expansion velocity")


p2_2 <- modsel_diversity  %>% 
  ggplot()+
  geom_point(aes(is.evo,clm_pseudo_r2,
                         col=is.DDD),
                     size=3)+
  geom_line(aes(is.evo,clm_pseudo_r2,
                group=is.DDD,
                col=is.DDD))+
  geom_hline(yintercept=clm_pseudo_r2(model=mod_div_null),lty=2)+ ## the aicc of the mortality only model
  scale_y_continuous("pseudo-R^2")+
  labs(title="neutral diversity")

(p2_1 + p2_2 + plot_layout(guides="collect")) & scale_x_discrete("when are trait covariables sampled?") &
  labs(col="Density dependence?")  & theme_bw() &
  theme(axis.title.y=element_markdown(),
        legend.text = element_markdown())
  
```

# figure 3: analysis of effects on velocities


# figure 4: analyses of effects on diersity


# fig5: analyse evolution

our main tool: the plot y-x~x where y=trait after evol, x=trait at t0, deviation from nthe line y-x=0 indicate evolution


```{r fig6}
p11<- tab %>% 
  filter(is.VP_dmax==1 & is.VP_DDD==1) %>% 
  ggplot()+
    geom_ribbon(data=tibble(x=c(-100:100)/100),
                aes(x=x,ymin=1-x,ymax=1),
                fill="darkgrey",col="black",alpha=0.5)+
    geom_ribbon(data=tibble(x=c(-100:100)/100),
                aes(x=x,ymin=-1-x,ymax=-1),
                fill="darkgrey",col="black",alpha=0.5)+
  # denotes the strategies possibles (davg>dmax not possible)
    geom_point(aes(start_avgslope,
                   mean_avgslope_100-start_avgslope),
               col="grey80")+
    geom_smooth(aes(start_avgslope,
                    mean_avgslope_100-start_avgslope),
                method="lm",
                col="black",fill="red")+
  labs(title='C: Individual variation in *&alpha;*, *&beta;*, and *d<sub>max</sub>*')+
  scale_y_continuous("")+
  scale_x_continuous("")

p10<- tab %>% 
  filter(is.VP_dmax==1 & is.VP_DDD==0) %>% 
  ggplot()+
    geom_ribbon(data=tibble(x=c(-100:100)/100),
                aes(x=x,ymin=1-x,ymax=1),
                fill="darkgrey",col="black",alpha=0.5)+
    geom_ribbon(data=tibble(x=c(-100:100)/100),
                aes(x=x,ymin=-1-x,ymax=-1),
                fill="darkgrey",col="black",alpha=0.5)+
  # denotes the strategies possibles (davg>dmax not possible)
    geom_point(aes(start_avgslope,
                   mean_avgslope_100-start_avgslope),
               col="grey80")+
    geom_smooth(aes(start_avgslope,
                    mean_avgslope_100-start_avgslope),
                method="lm",
                col="black",fill="red")+
 labs(title='B: Individual variation in *d<sub>max</sub>*')+
  scale_x_continuous("Density-dependence (*&Delta;<sub>avg-0</sub>*) at *t* = 0")+
  scale_y_continuous("")
#the white area delimitate the maximal space available for evolution (slope or davg after avolution cannot go higher than abs(maximal dmax possible under the variance conditions))
#note that no var vignettes should probably be all grey (no evolution possible)

p01<- tab %>% 
  filter(is.VP_dmax==0 & is.VP_DDD==1) %>% 
  ggplot()+
    geom_ribbon(data=tibble(x=c(-100:100)/100),
                aes(x=x,ymin=0.5-x,ymax=1),
                fill="darkgrey",col="black",alpha=0.5)+
    geom_ribbon(data=tibble(x=c(-100:100)/100),
                aes(x=x,ymin=-0.5-x,ymax=-1),
                fill="darkgrey",col="black",alpha=0.5)+
  # denotes the strategies possibles (davg>dmax not possible)
    geom_point(aes(start_avgslope,
                   mean_avgslope_100-start_avgslope),
               col="grey80")+
    geom_smooth(aes(start_avgslope,
                    mean_avgslope_100-start_avgslope),
                method="lm",
                col="black",fill="red")+
  scale_y_continuous("Change in *&Delta;<sub>avg-0</sub>* at the edge between *t* = 0 et *t* = 100")+
  scale_x_continuous("")+
  ggtitle("A: Individual variation in *&alpha;* and *&beta;*")

(p01+p10+p11) &
  geom_hline(yintercept = 0,lty=2) &
  coord_cartesian(xlim=c(-0.3,0.3),ylim=c(-0.25,0.8))  & 
  facet_grid(rows=vars(paste("*m* =",dispersal_mortality)),
               cols=vars(paste("*h^2* =",heritability))) &
  theme_bw()&
  theme(strip.text.x=element_markdown(),
        strip.text.y=element_markdown(),
        axis.title.x = element_markdown(),
        axis.title.y=element_markdown(),
        plot.title=element_markdown())

```





```{r fig5}
tab %>% 
  filter(is.VP_dmax==1) %>% 
  mutate(is.VP_DDD=factor(is.VP_DDD)) %>% 
  ggplot()+
    geom_hline(yintercept = 0.5,lty=2)+
    geom_jitter(aes(is.VP_DDD,mean_dmax_100),
               col="grey80")+
    stat_summary(aes(is.VP_DDD,mean_dmax_100),
                 fun = "mean",fun.min="mean",fun.max="mean",size=1)+
  scale_x_discrete("sources of individual variation",
                   labels=c("*d<sub>max</sub>* only",
                            "*&alpha;*, *&beta;*, and *d<sub>max</sub>*"))+
  scale_y_continuous("*d<sub>max</sub>* at *t* = 100")+
  facet_grid(rows=vars(paste("*m* =",dispersal_mortality)),
               cols=vars(paste("*h^2* =",heritability))) +
  theme_bw()+
  theme(legend.position="none",
        strip.text.x=element_markdown(),
        strip.text.y=element_markdown(),
        axis.text.x = element_markdown(),
        axis.title.y=element_markdown(),
        plot.title=element_markdown())
```




```{r}

newdata<-tibble(mean_avgslope_100=0,
                mean_d0_100=c(0:100)/100) %>%
  expand_grid(mortality=c("0.1","0.5","0.9")) %>% 
  mutate(row=1:length(mortality))

p1<-fitted_samples(mod_100,n=1000,newdata=newdata,seed=42,scale="response") %>% 
  left_join(newdata) %>% 
  ggplot()+
  ggdist::stat_lineribbon(aes(mean_d0_100,fitted),.width=c(0.001,0.95),fill="grey")+
  scale_y_continuous("predicted expansion velocity (patches/generation)")+
  scale_x_continuous("*d<sub>0</sub>* at *t* = 100")+
  facet_grid(rows=vars(paste("*m* = ", mortality)),
                       cols=vars("effect of *d<sub>0</sub>*"))+
  coord_cartesian(ylim=c(0,1))

newdata<-tibble(mean_avgslope_100=seq(
  from=range(tab$mean_avgslope_100)[1],
  to=range(tab$mean_avgslope_100)[2],
  length.out=100),
                mean_d0_100=mean(tab$mean_d0_100)) %>%
  expand_grid(mortality=c("0.1","0.5","0.9")) %>% 
  mutate(row=1:length(mortality))

p2<-fitted_samples(mod_100,n=1000,newdata=newdata,seed=42,scale="response") %>% 
  left_join(newdata) %>% 
  ggplot()+
  ggdist::stat_lineribbon(aes(mean_avgslope_100,fitted),.width=c(0.001,0.95),fill="grey")+
  scale_y_continuous("")+
  scale_x_continuous("*&Delta;<sub>avg-0</sub>* at *t* = 100")+
    facet_grid(rows=vars(paste("*m* = ", mortality)),
                       cols=vars("effect of density-dependence *&Delta;<sub>avg-0</sub>*"))+
  coord_cartesian(ylim=c(0,1))

(p1|p2)&
  theme_bw()&
  theme(axis.title.x = element_markdown(),
        strip.text.x = element_markdown(),
        strip.text.y = element_markdown())

appraise(mod_100)[[4]]+geom_abline(intercept=0,slope=1)
```


```{r}
newdata<-tibble(start_avgslope=0,
                start_d0=c(0:50)/100) %>%
  expand_grid(mortality=c("0.1","0.5","0.9")) %>% 
  mutate(row=1:length(mortality))

p1<-cbind(newdata,predict(mod_div_00,newdata=newdata)$fit) %>% 
  pivot_longer(cols=c(`20`,`40`,`60`,`80`,`100`,`120`,`121`)) %>% 
  mutate(name=fct_recode(name,         
                                       `< 20`="20",
                                       `21 to 40`="40",
                                       `41 to 60`="60",
                                       `61 to 80`="80",
                                       `81 to 100`="100",
                                       `101 to 120`="120",
                                       `> 120`="121")
  ) %>% 
  mutate(name=str_replace(as.character(name), "< 20", "\u2264 20" )) %>% 
  mutate(name=fct_relevel(name,
                                        "\u2264 20",
                                        "21 to 40",
                                        "41 to 60",
                                        "61 to 80",
                                        "81 to 100",
                                        "101 to 120",
                                        "> 120")) %>% 
  ggplot()+
  geom_area(aes(x=start_d0,y=value,fill=name),position="fill")+ 
  scale_y_continuous("")+
  scale_x_continuous("*d<sub>0</sub>* at *t* = 0")+
  guides(fill="none")+
  facet_grid(rows=vars(paste("*m* = ", mortality)),
                       cols=vars("effect of *d<sub>0</sub>*"))+
  coord_cartesian(ylim=c(0,1))

newdata<-tibble(start_avgslope=seq(
  from=range(tab$start_avgslope)[1],
  to=range(tab$start_avgslope)[2],
  length.out=100),
                start_d0=mean(tab$start_d0)) %>%
  expand_grid(mortality=c("0.1","0.5","0.9")) %>% 
  mutate(row=1:length(mortality))

p2<-cbind(newdata,predict(mod_div_00,newdata=newdata)$fit) %>% 
  pivot_longer(cols=c(`20`,`40`,`60`,`80`,`100`,`120`,`121`)) %>% 
  mutate(name=fct_recode(name,         
                                       `< 20`="20",
                                       `21 to 40`="40",
                                       `41 to 60`="60",
                                       `61 to 80`="80",
                                       `81 to 100`="100",
                                       `101 to 120`="120",
                                       `> 120`="121")
  ) %>% 
  mutate(name=str_replace(as.character(name), "< 20", "\u2264 20" )) %>% 
  mutate(name=fct_relevel(name,
                                        "\u2264 20",
                                        "21 to 40",
                                        "41 to 60",
                                        "61 to 80",
                                        "81 to 100",
                                        "101 to 120",
                                        "> 120")) %>% 
  ggplot()+
  geom_area(aes(x=start_avgslope,y=value,fill=name),position="fill")+ 
  scale_y_continuous("")+
  scale_x_continuous("*&Delta;<sub>avg-0</sub>* at *t* = 0")+
  labs(fill="generations before allele fixation:")+
  facet_grid(rows=vars(paste("*m* = ", mortality)),
                       cols=vars("effect of density-dependence *&Delta;<sub>avg-0</sub>*"))+
  coord_cartesian(ylim=c(0,1))



(p1|p2)&
  theme_bw() &
  scale_fill_brewer(palette="Greys") &
  plot_layout(guides="collect")&
  theme(axis.title.x = element_markdown(),
        strip.text.x = element_markdown(),
        strip.text.y = element_markdown())
```


###

to move to supplementary

## velocity evidence

patches are discrete entities, and speed is capped at one patch/generation. So let's start with binomial models:


```{r}

vf_dd=function(d0,f0){ # version for discrete space and time in Wang et al 2019
  ### 10.1016/j.tpb.2019.04.003
  
if(length(d0)!=length(f0)){stop("f0 and d0 must have same length")}
    
v=function(K,d,fec){
  return(log(fec*(1+d*(cosh(K)-1)))/K)
}

oo=NA
for(i in 1:length(d0)){
opti=optimize(f=v,d=d0[i],fec=f0[i],lower=0,upper=700)

#if brm (to avoid speeds >=1)
#if(opti$objective>0.9999){oo[i]=0.9999}
#else{oo[i]=opti$objective}

oo[i]=opti$objective
}
return(oo)
}

###we could use that vf in a proper brm, if we account for sources of stochasticity reducing
# true vf compare to nominal vf
#mod<-brm(bf(advance|trials(20)~logit(vf3)+d,
#         d~(mean_d0_100+mean_avgslope_100)*mortality+is.VP_DDD+is.VP_dmax,
#         nl=TRUE,family=binomial),
#         data=subset(tab2,advance>-1),
#         prior=c(set_prior("normal(0,1)",nlpar="d",class="b")),
#         chains=4,iter=200,backend="cmdstanr")



tab2 <-tab %>% 
  mutate(d0_eff=(1-dispersal_mortality)*mean_d0_100,
         r0_eff=log(5-(dispersal_mortality*d0_eff)*5),
         vf=2*(log(5)*0.5*d0_eff)^0.5,
         vf2=2*(r0_eff*0.5*d0_eff)^0.5) %>% 
  mutate(vf3=vf_dd(d0_eff,exp(r0_eff))) %>% 
  arrange(d0_eff)

tab2 %>% 
  filter(mean_avgslope_100>-0.005 & mean_avgslope_100<0.005) %>% 
  ggplot()+ #only slope close to 0
  geom_point(aes(d0_eff, ##use effective d0, accounting for motality
                 advance/20,col=mortality))+
  geom_line(aes(d0_eff,vf2))+
  geom_line(aes(d0_eff,vf3))+
  geom_line(aes(d0_eff,vf),lty=2)+
  facet_wrap(~dispersal_mortality+heritability)


tab2 %>% 
  ggplot()+ #only slope close to 0
  geom_point(aes(mean_avgslope_100, ##use effective d0, accounting for motality
                 (advance/20)/vf3,col=mortality))+
  facet_wrap(~dispersal_mortality)

## shows that the continuum approx for v is goodish until it bumps into the v=1 boundary
## birzu says approach is very good when r0<<0 and d0/r0 >>1
## we're not in these conditions but still good enough

## we won't use our models to try and disentangle effect of mortality on effective d0 versus
## other effects, but nice to know it exists
```



```{r}
tab %>% 
  mutate(round_d0=round(start_d0,1)) %>% 
  filter(start_avgslope <0.05 & start_avgslope>-0.05) %>% 
  group_by(round_d0,mortality,fixinterval) %>% 
  count() %>% 
  ggplot()+
  geom_col(aes(x=round_d0,y=n,fill=fixinterval),position=position_fill())+
  facet_grid(rows=vars(mortality))


tab %>% 
  mutate(round_slope=round(start_avgslope,2)) %>% 
  #filter(start_avgslope <0.05 & start_avgslope>-0.05) %>% 
  group_by(round_slope,mortality,fixinterval) %>% 
  count() %>% 
  ggplot()+
  geom_col(aes(x=round_slope,y=n,fill=fixinterval),position=position_fill())+
  facet_grid(rows=vars(mortality))

```

