---
title: 'main analysis code for "Individual variation in dispersal, and its sources, shape the fate of pushed vs. pulled range expansions"'
author: "Maxime Dahirel, Chloé Guicharnaud, Elodie Vercken (code by Maxime Dahirel)"
date:
output:
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

```{r load-packages}
library(tidyverse)

library(brms)
library(bayesplot)
library(tidybayes)
options(mc.cores=4)

library(patchwork)   # CRAN v1.1.1
library(ggtext)

library(here)        # CRAN v1.0.1
```

# Introduction

After running the code in `R/generate_simulations`, we obtained, and saved, a `csv` file containing patch-level data for every simulation we ran. We can now load it, but we're going to immediately make two quick additions: 
- in the IBM, phenotypic variance input variables denote, well, the actual phenotypic variances we wanted to input. But since we only tested two levels, variance or no variance, much easier for us to work with binary variables going forward
- we create a `scenario` variable that groups together all replicates that share the same initial conditions (fecundity, mortality, starting values of dispersal traits, their variability and heritability).

```{r import-data}
data <- read_csv(here("NetLogo_output", "simulation_output.csv")) %>%
  mutate(
    is.VP_dmax = as.numeric(VP_logit_dmax > 0),
    is.VP_DDD = as.numeric(VP_slope > 0)
    # if VP_slope >0, then VP_midpoint is also >0 in our experimental design
    # so only one binary variable needed for density-dependent dispersal variables
  )|>
  mutate(scenario=paste0("F=",fecundity, "_M=",dispersal_mortality,
                         "_h2=",heritability,
                     "_slope=",start_slope,"_midpoint=",start_midpoint,
                     "_VP.DDD=",is.VP_DDD,"_VP_dmax=",is.VP_dmax)) 
```

OK, the dataset we just imported is structured so that one row = one patch in one landscape at one observed time. Only patches that were populated at least once are included (see the end of the `generate_simulations` file). The dataset contains many variables, many of them were collected as a precaution "in case they were needed later" (since re-running the models is time consuming); not all of them will be used in the analysis:

- `ticks`: time step at which the observations were made (= number of generations since start), we planned to collect data only at the end of the runs (`ticks` = 121). As `nlrx` may fail if no data are collected for a run, we also collected data at `ticks` = 1, to avoid issues with replicates that go extinct before the end of their run.

Landscape-level variables, all constant across time (either because they are truly constant, or because they only describe initial conditions at $t=0$ and are not updated after):
              
- `start_dmax`, `start_slope`, `start_midpoint`: mean dispersal trait values at the start of the experiment (`slope` corresponds to $\alpha$ in the manuscript, `midpoint` to $\beta$)
- `VP_logit_dmax`, `VP_slope`, `VP_midpoint`: **initial** phenotypic variances
- `is.VP_dmax`, `is.VP_DDD`: binary variables describing whether `VP_logit_dmax` on one hand, and `VP_slope` and `VP_midpoint` on the other, are >0
- `heritability`: **initial** heritability (same value for all traits)
- `dispersal_mortality`: probability of dying during dispersal
- `fecundity`: hypothetical average fecundity at $N=0$ ($\exp(r_{0})$)   
- `scenario` : combination of `fecundity`, `dispersal_mortality`, `heritability`, `start_slope`, `start_midpoint`, `is.VP_dmax`, `is.VP_DDD`. Summarises the initial state of the expansion
- `K`: patch carrying capacity (same for all patches)
- `reproduction`: whether the simulated species was clonal or sexual (here all runs are clonal)
- `seedID`, `siminputrow`, `replicateID`: the first two variables, combined, give the third one: a unique identifier for each landscape/run
- `start_d0`, `start_dK`, `start_slopeA_0_K`: mean values for $d_{0}$, $d_{K}$, + the density-dependence metric $\Delta_{K-0}$ 

Patch-level variables:
- `pxcor`: patch $x$ coordinate                   
- `founding`: the generation at which the patch was first populated
- `N_predispersal` and `N_postdispersal`: population size in the patch immediately before and after the dispersal phase, respectively
- `N_sedentary`: number of individuals born in the patch that stayed in it
- `N_disp_dead`: number of individuals born in the patch that died during dispersal             
- `N_allele0_pre`, `N_allele1_pre`, `N_allele0_post`, `N_allele1_post`: number of individuals with each neutral allele, pre- and post-dispersal
- `mean_genotype_logit_dmax`, `var_genotype_logit_dmax`, `mean_genotype_slope`, `var_genotype_slope`, `mean_genotype_midpoint`,  `var_genotype_midpoint`: mean and variances  of the genetic component of traits   
- `mean_dmax`, `mean_slope`, `mean_midpoint`: phenotypic means of the dispersal traits
- `mean_d0`, `mean_dK`, `mean_slopeA_0_K`: means of dispersal metrics that can be estimated from these dispersal traits (see above)     

# Data wrangling

To do our analyses, we're going to build several "processed" tables from this "raw" dataset, which we'll then merge into a final table.

Again: we will only focus on the state of our expansions at the end of the simulations (t = 121, effectively after 120 generations of expansion; see below as to why). There is good evidence that evolutionary dynamics are at or close to equilibrium for most replicates at this time. (see Supplementary)

Second: we will consistently use the metrics as captured at the *start* of each generation, i.e. pre-dispersal. This is to avoid trying to interpret non-genetic dynamics in traits that are reset each generation. For instance, if dispersal capacity is variable but 100% non-genetic, there may be an excess of high dispersers in newly colonised patches. This excess will however vanish when they reproduce and dispersal traits are redrawn. In front patches that are funded by a low number of individuals, this also allows us to estimate patch averages for traits and genetic diversity from higher post-reproduction numbers vs. lower founding numbers.

We start by creating a backbone data frame with one row for **each** replicate, extinct or not

```{r backbone}

tab <- data |>  
  filter(ticks ==1) |> 
  select(replicateID, scenario, 
           dispersal_mortality, heritability, fecundity,
           K,
           is.VP_dmax, is.VP_DDD,
           start_slope, start_midpoint,
           start_d0, start_dK,        
           start_slopeA_0_K) |> 
  distinct()

```


We then create a dataset for genetic diversity at the end of the expansion, with one observation = one replicate expansion *that expanded past x = 0*. Our metric of genetic diversity is spatial, and based on the positions of the expansion front for the whole expansion vs. for the individuals that bear the neutral allele that lags behind. 

```{r genet_data}
tab_genet <- data  |>
  filter(ticks==121) |> # we only keep the last iteration
  filter(N_predispersal>0) |> # for genetic diversity to be measured in a patch, there needs to be individuals
  group_by(replicateID) |> 
  summarise(edge1 = max(pxcor * (N_allele1_pre > 0), na.rm=TRUE), # the coordinate of the farthest individual with neutral allele 1
            edge0 = max(pxcor * (N_allele0_pre > 0), na.rm=TRUE)  # same with neutral allele 0
  ) |> 
  rowwise() |> 
  mutate(edge_pre = max(edge0, edge1),  # which is the farthest of the two irrespective of allele value
         lagging_pre = min(edge0, edge1) # which is the other?
  ) |> 
  ungroup()
```

(the pre-dispersal state at generation 121 reflects what happened after 120 generations of expansion)

We then create a dataset that contain the mean dispersal trait values at time $t$ in the range front (which we defined here semi-arbitrarily as patches < 5 patches from the edge; see Supplementary S2 for justification).

We first use the previously created dataset to know, for each replicate, which patches are "front patches" (we use the edge pre-dispersal as a reference, since traits are measured pre-dispersal)

```{r tab_intermediate}
tab_intermediate <- tab_genet |> 
  select(replicateID, edge_pre)

```

We then use that to (i) filter the big `data` to only keep front patches (ii) and then get weighted averages for each key trait 


```{r make_tab_traits}
tab_traits <- data |> 
  filter(ticks==121) |> 
  left_join(tab_intermediate) |> 
  filter(pxcor > (edge_pre - 5)) |> 
  group_by(replicateID)  |> 
  summarise(
    mean_dmax = weighted.mean(mean_dmax, w = N_predispersal, na.rm = TRUE),
    mean_dK = weighted.mean(mean_dK, w = N_predispersal, na.rm = TRUE),
    mean_d0 = weighted.mean(mean_d0, w = N_predispersal, na.rm = TRUE),
    mean_Kslope = weighted.mean(mean_slopeA_0_K, w = N_predispersal, na.rm = TRUE)
  )  |> 
  ungroup()
```

We can then merge the two in one dataset, and we rename one variables so that models are a bit easier to write below. We also add a check to distinguish which expansions are successfull, stalled at x = 0 or went fully extinct

```{r combined_tab}
tab <- left_join(tab,tab_genet) |> 
  left_join(tab_traits) |> 
  mutate(Mort=factor(dispersal_mortality),
         Fec=factor(fecundity)
         ) |> 
  rename(start_Kslope="start_slopeA_0_K") |> 
  mutate(status = case_when(is.na(edge_pre)==TRUE ~ "extinct",
                            edge_pre==0 ~ "failed",
                            TRUE ~ "expanded"))
```

```{r}
table(tab$status,tab$Mort,tab$Fec)
# all failed or extinct are in low-fecundity, high-mortality, no exceptions
```

For the analysis of expansions velocities, we might want to count the failed expansions as 0s, not NAs. So let's create a final variable for that:

```{r}
tab$obs_v <- case_when(tab$status=="extinct" ~ 0,
                     TRUE ~ tab$edge_pre)
```


# Analysis

## Analysing expansion velocities, part 1: preparation

We first implement a function that takes per capita fecundity (*not growth rate!*) and dispersal rate, and gives the expected velocity $v^*$ in the absence of dispersal density-dependence. We use the formula in Wang et al. 2019 (doi: 10.1016/j.tpb.2019.04.003) that accounts for discrete time and discrete space (with dispersal to nearest neighbour only); otherwise we might wrongly predict velocities > 1  when they are not possible in that setup. We modify the formula to account for dispersal mortality: dispersal mortality reduces both the net dispersal rate and the net fecundity that effectively enter the Wang et al formula

```{r vstar_wang19}
vstar <- function(fec,disp, mort, na.low = FALSE){
  
  # tests
  if(!is.logical(na.low) | is.na(na.low)){stop("na.low must be TRUE or FALSE!")}
  if(!is.vector(fec) | !is.vector(disp)| !is.vector(mort)){stop("all inputs should be numeric vectors!")}
  if(length(fec)!= 1 | length(disp)!=1 | length(mort)!=1){stop("all inputs should be of size 1!")}
  if(is.na(fec) | is.na(disp) |is.na(mort)){
    warning("one of the arguments is NA, will return NA")
    return( NA_real_ )
    }
  if(!is.numeric(fec) | !is.numeric(disp) | !is.numeric(mort)){stop("all inputs should be numeric vectors!")}
  
  if(fec < 0){stop("fec should be >= 0! (it is per-capita fecundity, not growth rate)")}
  if(disp < 0 | disp > 1){stop("dispersal rate disp should be between 0 and 1!")}
  if(mort < 0 | mort > 1){stop("dispersal mortality mort should be between 0 and 1!")}
  
  # formula in Wang et al Appendix A1
  test = optimise(
  function(kappa,
           rho=fec*(1-disp*mort), # individuals dying in dispersal are lost to net fecundity
           m=disp*(1-mort)        # and also obviously to net dispersal
           ){
    log(rho*(1+ m *(cosh(kappa)-1)))/kappa
    },
    lower=0,upper=707 #cosh(kappa>709) = overflow, estimated as Inf; upper a few orders lower by safety
)
  # Possibly due to the numerical optimisation, 
  # the above will actually plateau a few 10^-3 or 10^-4 above 1, rather than 1
  # It will also give negative values at fecundities < 1
  # so some post-processing needed
  vstar <- test$objective
  if(vstar >1){vstar <- 1}
  if(vstar < 0 & na.low == FALSE){vstar <- 0}
  if(vstar < 0 & na.low == TRUE){vstar <- NA_real_}
return(vstar)
}
```

We can apply the function to our data with the help of `map`:

```{r create_vstar}
tab <- tab |> 
  group_by(replicateID) |> 
  nest() |> 
  mutate(vstar = map(
    .x=data,
    .f = \(x) vstar(
    fec=x$fecundity,
    disp=x$start_d0,
    mort=x$dispersal_mortality)
    )
    ) |> 
  mutate(vstar_final = map( #ignore the warning in vstar_final, caused by extinctions
    .x=data,
    .f = \(x) vstar(
    fec=x$fecundity,
    disp=x$mean_d0,
    mort=x$dispersal_mortality)
    )
    )|> 
  unnest(data) |> 
  unnest(vstar) |> 
  unnest(vstar_final) |> 
  ungroup()
```

## Splitting the data

We can't really combine in one analysis all the simulations: (i) expansions without phenotypic variation, (ii) with only non-heritable phenotypic variation and (iii) where evolution is possible in at least one trait are three "regimes" that are best analysed separately.

So let's split the data:

```{r split_tab}

tab_fix = filter(tab, is.VP_DDD ==0 & is.VP_dmax ==0) #(i) traits are fixed

tab_noevo = filter(tab,(is.VP_DDD==1 | is.VP_dmax==1) & heritability ==0) #(ii) traits can vary but do not evolve
tab_evo = filter(tab,(is.VP_DDD==1 | is.VP_dmax==1) & heritability == 1) #(iii) at least one trait can evolve

dim(tab)[1] == (dim(tab_fix)[1] + dim(tab_noevo) [1] + dim(tab_evo)[1]) #sense check

```

For (i) and (ii) there is no reason to analyse phenotypic trait changes, this can only be done meaningfully in (iii) (and even in iii, not all trait changes can be analysed for all replicates, as we'll see later)

## Analysing expansion velocities, part 2

```{r cor_tests}
cor.test(tab_fix$vstar,tab_fix$obs_v/120)
cor.test(tab_noevo$vstar,tab_noevo$obs_v/120)


cor.test(tab_evo$vstar,tab_evo$obs_v/120)
cor.test(tab_evo$vstar_final,tab_evo$obs_v/120)

#lmer_fix=lmer(obs_v/120~vstar+(1|scenario),data=tab_fix)
#lmer_noevo=lmer(obs_v/120~vstar+(1|scenario),data=tab_noevo)
#lmer_evo1=lmer(obs_v/120~vstar+(1|scenario),data=tab_evo)
#lmer_evo2=lmer(obs_v/120~vstar_final+(1|scenario),data=tab_noevo)
```

Let's make Figure 3:

```{r fig3_a}
fig3_a <- ggplot(tab_fix)+
  geom_point(aes(vstar,obs_v/120,fill=start_Kslope),pch=21,size=2.5)+
  geom_abline(intercept = 0,slope=1)+
  #geom_abline(intercept = fixef(lmer_fix)[1],slope=fixef(lmer_fix)[2],linetype=2)+
  scale_fill_viridis_c("&Delta;<sub>K-0</sub>", option = "inferno")+
  scale_x_continuous("")+
  scale_y_continuous("Observed velocity")+
  ggtitle("A) no trait variation") +
  theme_bw()+
  theme(
    legend.position = "none", 
    # legends of 3_a and 3_b are identical (you can check by restoring this legend)
    # but plot_layout(guides="collect") does not recognise this and leaves the duplicate
    axis.title.x = element_markdown()
    )
```

```{r fig3_b}
fig3_b <- ggplot(tab_noevo)+
  geom_point(aes(vstar,obs_v/120,fill=start_Kslope),pch=21,size=2.5)+
  geom_abline(intercept = 0,slope=1)+
  #geom_abline(intercept = fixef(lmer_noevo)[1],slope=fixef(lmer_noevo)[2],linetype=2)+
  scale_fill_viridis_c("&Delta;<sub>K-0</sub>", option = "inferno")+
  scale_x_continuous("Theoretical velocity if density-independent dispersal (i.e. &Delta;<sub>K-0</sub> = 0)")+
  scale_y_continuous("Observed velocity")+
  ggtitle("B) trait variation but no evolution")+
  theme_bw()+
  theme(
    legend.title=element_markdown(),
    axis.title.x = element_markdown()
    )
```

```{r fig3}
(fig3_a/fig3_b) + plot_layout(guides="collect")
```

And then Figure 4:

```{r fig4_a}
fig4_a <- ggplot(tab_evo)+
  geom_point(aes(vstar,obs_v/120,fill=start_Kslope),pch=21,size=2.5)+
  geom_abline(intercept = 0,slope=1)+
  #geom_abline(intercept = fixef(lmer_evo1)[1],slope=fixef(lmer_evo1)[2],linetype=2)+
  scale_fill_viridis_c("**Initial** &Delta;<sub>K-0</sub>", option = "inferno")+
  scale_x_continuous("")+
  scale_y_continuous("Observed velocity")+
  ggtitle("A) evolving replicates, using initial <i>v^*^</i> and &Delta;<sub>K-0</sub>") +
  theme_bw()+
  theme(
    plot.title=element_markdown(),
    legend.title=element_markdown(),
    axis.title.x = element_markdown()
    )
```

```{r fig4_b}
fig4_b <- ggplot(tab_evo)+
  geom_point(aes(vstar_final,obs_v/120,fill=mean_Kslope),pch=21,size=2.5)+
  geom_abline(intercept = 0,slope=1)+
  #geom_abline(intercept = fixef(lmer_evo2)[1],slope=fixef(lmer_evo2)[2],linetype=2)+
  scale_fill_viridis_c("**Final** &Delta;<sub>K-0</sub>", option = "inferno")+
  scale_x_continuous("Theoretical velocity if density-independent dispersal (i.e. &Delta;<sub>K-0</sub> = 0)")+
  scale_y_continuous("Observed velocity")+
  ggtitle("B) evolving replicates, using final <i>v^*^</i> and &Delta;<sub>K-0</sub> (_t_=120)") +
  theme_bw()+
  theme(
    plot.title=element_markdown(),
    legend.title=element_markdown(),
    axis.title.x = element_markdown()
    )
```


```{r fig4}
(fig4_a/fig4_b)
```


## Priors for both genetic diversity and trait evolution models

```{r model_priors}
prior_binomial <- c(
                    set_prior("normal(0,1.5)",class="Intercept"),
                    set_prior("normal(0,1)",class="b"),
                    set_prior("normal(0,1)",class="sd")
                  )

prior_beta_binomial <- c(
                    set_prior("normal(0,1.5)",class="Intercept"),
                    set_prior("normal(0,1)",class="b"),
                    set_prior("normal(0,1)",class="sd"),
                    set_prior("normal(0,1)",nlpar="loginvphi",class="b")
                  )

prior_beta <- prior_beta_binomial
```


## Genetic diversity models

```{r}
mod_gen_fixB <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + 
                          (1|scenario)),
                  family=binomial,
                  data=subset(tab_fix,status=="expanded"), #excludes extinct and non-advancing expansions
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior= prior_binomial
)

pp_check(mod_gen_fixB,"stat_2d") #some evidence of overdispersion?

mod_gen_fixBB0 <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + (1|scenario),
                         nlf(phi~1/exp(loginvphi)),
                         # the "default" overdispersion parameter in brms is actually a precision parameter
                         # the closest to Inf the closest to binomial
                         # we want an overdispersion parameter (as in Harrison 2015), so we inverse
                         loginvphi~1),
                  family=beta_binomial(link_phi="identity"), # we need identity link so the transform works well
                  data=subset(tab_fix,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_fixBB0,"stat_2d") # better

mod_gen_fixBB <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ start_Kslope * Mort * Fec  + (1|scenario)),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_fix,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_fixBB,"stat_2d") # good

loo_compare(
  loo(mod_gen_fixB),
  loo(mod_gen_fixBB0),
  loo(mod_gen_fixBB)
)
## even with a few "Pareto k problematic obs", the differences in loo and the pp_check are clear
```


```{r}
select(tab_fix,Mort,Fec,scenario) |> distinct() |> group_by(Mort,Fec) |> count() 
#210 scenarios possible, so in theory 5*210 replicates = 1050
#7 scenarios went fully extinct/failed to advance, so 203, and a max of 1015 replicates
# we have 1009, so some scenarios failed elsewhere
```

```{r}
fits_fix=select(tab_fix,Mort,Fec) |> distinct() |> 
  mutate(edge_pre=1) |> 
  expand_grid(start_Kslope=c(-5:5)/10) |> 
  add_epred_draws(mod_gen_fixBB,re_formula=NA) |> 
  ungroup()

tab_fix |>  
  ggplot()+
  geom_point(aes(x=start_Kslope,y=lagging_pre/edge_pre,size=edge_pre),col="lightblue",alpha=0.25)+
  stat_lineribbon(data=fits_fix,aes(x=start_Kslope,y=.epred),fill="grey",col="grey40")+
  geom_point(data=tab_fix |> group_by(scenario,Mort,Fec) |> 
               summarise(div=mean(lagging_pre/edge_pre,na.rm=TRUE),
                         slope=mean(start_Kslope)),
             aes(x=slope,div),size=2,fill="white",pch=21)+
  facet_grid(rows=vars(paste0("*r<sub>0</sub>* = log(",Fec,")")),cols=vars(paste("*m* =",Mort)))+
  scale_x_continuous("initial dispersal density-dependence (*&Delta;<sub>K-0</sub>*)")+
  scale_y_continuous("how far does genetic diversity persist?")+
  ggtitle(label="",
          subtitle="A) replicates with no individual variation in dispersal traits")+
  theme_bw()+
  theme(legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(angle = 0),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.subtitle = element_markdown()
  )
```

```{r}
mod_gen_noevoB <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + 
                          (1|scenario)),
                  family=binomial,
                  data=subset(tab_noevo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_binomial
)

pp_check(mod_gen_noevoB,"stat_2d")

mod_gen_noevoBB0 <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + 
                          (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ 1),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_noevo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_noevoBB0,"stat_2d")

mod_gen_noevoBB <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ start_Kslope * Mort * Fec  + (1|scenario)),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_noevo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_noevoBB,"stat_2d") # good

loo_compare(
  loo(mod_gen_noevoB),
  loo(mod_gen_noevoBB0),
  loo(mod_gen_noevoBB)
)
```


```{r}
select(tab_noevo,Mort,Fec,scenario) |> distinct() |> group_by(Mort,Fec) |> count() 
#630 scenarios possible, so in theory 5*630 replicates = 3050
#9 scenarios went fully extinct/failed to advance (again all at max mortality low fec), so 621, and a max of 3105 replicates
# we have 3092, so some scenarios failed elsewhere
```

```{r}
fits_noevo=select(tab_noevo,Mort,Fec) |> distinct() |> 
  mutate(edge_pre=1) |> 
  expand_grid(start_Kslope=c(-5:5)/10) |> 
  add_epred_draws(mod_gen_noevoBB,re_formula=NA) |> 
  ungroup()

outline_fix = fits_fix |> 
  ungroup() |> 
  group_by(Mort,Fec, start_Kslope) |> 
  mean_hdci(.epred) |> 
  ungroup()

tab_noevo |>  
  ggplot()+
  geom_point(aes(x=start_Kslope,y=lagging_pre/edge_pre,size=edge_pre),col="pink",alpha=0.25)+
  stat_lineribbon(data=fits_noevo,aes(x=start_Kslope,y=.epred),fill="grey",col="grey40")+
  geom_line(data=outline_fix,aes(x=start_Kslope,y=.lower),lty=2)+
  geom_line(data=outline_fix,aes(x=start_Kslope,y=.upper),lty=2)+
  geom_point(data=tab_noevo |> group_by(scenario,Mort,Fec) |> 
               summarise(div=mean(lagging_pre/edge_pre,na.rm=TRUE),
                         slope=mean(start_Kslope)),
             aes(x=slope,div),size=2,fill="white",pch=21)+
  facet_grid(rows=vars(paste0("*r<sub>0</sub>* = log(",Fec,")")),cols=vars(paste("*m* =",Mort)))+
  scale_x_continuous("initial dispersal density-dependence (*&Delta;<sub>K-0</sub>*)")+
  scale_y_continuous("how far does genetic diversity persist?")+
  ggtitle(label="",
          subtitle="B) individual variation in dispersal traits but no evolution (*h<sup>2</sup>* = 0)")+
  theme_bw()+
  theme(legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(angle = 0),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.subtitle = element_markdown()
  )
```


```{r}
mod_gen_evoB <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + 
                          (1|scenario)),
                  family=binomial,
                  data=subset(tab_evo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_binomial
)
# bad ESS (more than other binom models so far), probably linked to model itself being bad (more than other...)

pp_check(mod_gen_evoB,"stat_2d")

mod_gen_evoBB0 <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + 
                          (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ 1),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_evo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_evoBB0,"stat_2d")

mod_gen_evoBB <- brm(bf(lagging_pre|trials(edge_pre)~ start_Kslope * Mort * Fec  + (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ start_Kslope * Mort * Fec  + (1|scenario)),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_evo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_evoBB,"stat_2d") # good

loo_compare(
  loo(mod_gen_evoB),
  loo(mod_gen_evoBB0),
  loo(mod_gen_evoBB)
)
```


```{r}
fits_evo=select(tab_noevo,Mort,Fec) |> distinct() |> 
  mutate(edge_pre=1) |> 
  expand_grid(start_Kslope=c(-5:5)/10) |> 
  add_epred_draws(mod_gen_evoBB,re_formula=NA)

outline_noevo = fits_noevo |> 
  ungroup() |> 
  group_by(Mort,Fec, start_Kslope) |> 
  mean_hdci(.epred) |> 
  arrange(start_Kslope) |> 
  ungroup()

tab_evo |>  
  ggplot()+
  geom_point(aes(x=start_Kslope,y=lagging_pre/edge_pre,size=edge_pre),col="palegreen",alpha=0.25)+
  stat_lineribbon(data=fits_evo,aes(x=start_Kslope,y=.epred),fill="grey",col="grey40")+
  geom_line(data=outline_noevo,aes(x=start_Kslope,y=.lower),lty=2)+
  geom_line(data=outline_noevo,aes(x=start_Kslope,y=.upper),lty=2)+
  geom_point(data=tab_evo |> group_by(scenario,Mort,Fec) |> 
               summarise(div=mean(lagging_pre/edge_pre,na.rm=TRUE),
                         slope=mean(start_Kslope)),
             aes(x=slope,div),size=2,fill="white",pch=21)+
  facet_grid(rows=vars(paste0("*r<sub>0</sub>* = log(",Fec,")")),cols=vars(paste("*m* =",Mort)))+
  scale_x_continuous("initial dispersal density-dependence (*&Delta;<sub>K-0</sub>*)")+
  scale_y_continuous("how far does genetic diversity persist?")+
  ggtitle(label="",
          subtitle="C) individual variation in dispersal traits *and* evolution possible (*h<sup>2</sup>* = 1)")+
  theme_bw()+
  theme(legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(angle = 0),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.subtitle = element_markdown()
  )
```

Evo, but using traits post evo:

```{r}
mod_gen_postevoB <- brm(bf(lagging_pre|trials(edge_pre)~ mean_Kslope * Mort * Fec  + 
                          (1|scenario)),
                  family=binomial,
                  data=subset(tab_evo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_binomial
)

pp_check(mod_gen_postevoB,"stat_2d")

mod_gen_postevoBB0 <- brm(bf(lagging_pre|trials(edge_pre)~ mean_Kslope * Mort * Fec  + 
                          (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ 1),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_evo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_postevoBB0,"stat_2d")

mod_gen_postevoBB <- brm(bf(lagging_pre|trials(edge_pre)~ mean_Kslope * Mort * Fec  + (1|scenario),
                        nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ mean_Kslope * Mort * Fec  + (1|scenario)),
                  family=beta_binomial(link_phi="identity"),
                  data=subset(tab_evo,status=="expanded"),
                  backend="cmdstanr",
                  iter = 4000, warmup = 2000,
                  seed=42,
                  prior=prior_beta_binomial
)

pp_check(mod_gen_postevoBB,"stat_2d") # good

loo_compare(
  loo(mod_gen_postevoB),
  loo(mod_gen_postevoBB0),
  loo(mod_gen_postevoBB)
)
```


```{r}
fits_postevo=select(tab_evo,Mort,Fec) |> distinct() |> 
  mutate(edge_pre=1) |> 
  expand_grid(mean_Kslope=c(-10:10)/10) |> 
  add_epred_draws(mod_gen_postevoBB,re_formula=NA)

outline_evo = fits_evo |> 
  ungroup() |> 
  group_by(Mort,Fec, start_Kslope) |> 
  mean_hdci(.epred) |> 
  arrange(start_Kslope) |> 
  ungroup()

tab_evo |>  
  ggplot()+
  geom_point(aes(x=mean_Kslope,y=lagging_pre/edge_pre,size=edge_pre),col="palegreen",alpha=0.25)+
  stat_lineribbon(data=fits_postevo,aes(x=mean_Kslope,y=.epred),fill="grey",col="grey40")+
  #geom_line(data=outline_evo,aes(x=start_Kslope,y=.lower),lty=2)+
  #geom_line(data=outline_evo,aes(x=start_Kslope,y=.upper),lty=2)+
  geom_point(data=tab_evo |> group_by(scenario,Mort,Fec) |> 
               summarise(div=mean(lagging_pre/edge_pre,na.rm=TRUE),
                         slope=mean(mean_Kslope)),
             aes(x=slope,div),size=2,fill="white",pch=21)+
  facet_grid(rows=vars(paste0("*r<sub>0</sub>* = log(",Fec,")")),cols=vars(paste("*m* =",Mort)))+
  scale_x_continuous("**final** dispersal density-dependence (*&Delta;<sub>K-0</sub>*)")+
  scale_y_continuous("how far does genetic diversity persist?")+
  ggtitle(label="",
          subtitle="C) individual variation in dispersal traits *and* evolution possible (*h<sup>2</sup>* = 1)")+
  theme_bw()+
  theme(legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(angle = 0),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.subtitle = element_markdown()
  )
```




## evo models : slope

```{r}
tab_evo_DDD <- filter(tab_evo, is.VP_DDD==1 & status =="expanded") |>
  mutate(mean_Kslope_std = mean_Kslope/mean_dmax)
  


mod_slope <- brm(bf((mean_Kslope_std/2)+0.5~ start_Kslope * Mort * Fec  + (1|scenario),
                    nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ start_Kslope * Mort * Fec  + (1|scenario)),
                  family=Beta(link_phi="identity"),
                  data=tab_evo_DDD,
                 iter=4000, warmup =2000,
                 backend="cmdstanr",
                 seed=42,
                  prior=prior_beta
)
```

```{r}
fits_slope=select(tab_evo_DDD,Mort,Fec) |> distinct() |> 
  expand_grid(start_Kslope=c(-5:5)/20) |> 
  add_epred_draws(mod_slope,re_formula=NA) |> 
  mutate(.epred = (.epred - 0.5) * 2) # rescaling to the -dmax/+dmax range

tab_evo_DDD |>  
  ggplot()+
  geom_abline(intercept = 0,slope=2,lty=2)+ 
  # ^ relative slope = absolute_slope / dmax  = absolute_slope / 0.5 at t=0
  # y = x/0.5 is y = x*2
  geom_point(aes(x=start_Kslope,y=mean_Kslope/mean_dmax),col="palegreen",alpha=0.25)+
  stat_lineribbon(data=fits_slope,aes(x=start_Kslope,y=.epred),fill="grey",col="grey40")+
  geom_point(data=tab_evo_DDD |> group_by(scenario,Mort,Fec) |> 
               summarise(trait=mean(mean_Kslope/mean_dmax,na.rm=TRUE),
                         slope=mean(start_Kslope)),
             aes(x=slope,trait),size=2,fill="white",pch=21)+
  facet_grid(rows=vars(paste0("*r<sub>0</sub>* = log(",Fec,")")),cols=vars(paste("*m* =",Mort)))+
  scale_x_continuous("initial dispersal density-dependence (*&Delta;<sub>K-0</sub>*)")+
  scale_y_continuous("relative dispersal density-dependence (*&Delta;<sub>K-0</sub> / d<sub>max</sub>*) after expansion")+
  theme_bw()+
  ggtitle(label="",subtitle="(only scenarios with heritable variation in *&alpha;* and *&beta;*)")+
  theme(legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(angle = 0),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.subtitle = element_markdown()
  )
```

## evo models: dmax

```{r}
tab_evo_dmax <- filter(tab_evo, is.VP_dmax==1  & status =="expanded")
  
mod_dmax <- brm(bf(mean_dmax~ start_Kslope * Mort * Fec  + (1|scenario),
                    nlf(phi~1/exp(loginvphi)),
                        loginvphi ~ start_Kslope * Mort * Fec  + (1|scenario)),
                  family=Beta(link_phi="identity"),
                  data=tab_evo_dmax,
                iter = 4000, warmup = 2000,
                backend="cmdstanr",
                seed=42,
                control = list(adapt_delta = 0.9), #some divergent transitions
                  prior=prior_beta
)
```

```{r}
fits_dmax=select(tab_evo_dmax,Mort,Fec) |> distinct() |> 
  expand_grid(start_Kslope=c(-5:5)/10) |> 
  add_epred_draws(mod_dmax,re_formula=NA)

tab_evo_dmax |>  
  ggplot()+
  geom_hline(yintercept = 0.5,lty=2) +
  geom_point(aes(x=start_Kslope,y=mean_dmax),col="palegreen",alpha=0.25)+
  stat_lineribbon(data=fits_dmax,aes(x=start_Kslope,y=.epred),fill="grey",col="grey40")+
  geom_point(data=tab_evo_dmax |> group_by(scenario,Mort,Fec) |> 
               summarise(trait=mean(mean_dmax,na.rm=TRUE),
                         slope=mean(start_Kslope)),
             aes(x=slope,trait),size=2,fill="white",pch=21)+
  facet_grid(rows=vars(paste0("*r<sub>0</sub>* = log(",Fec,")")),cols=vars(paste("*m* =",Mort)))+
  scale_x_continuous("initial dispersal density-dependence (*&Delta;<sub>K-0</sub>*)")+
  scale_y_continuous("dispersal capacity (*d<sub>max</sub>*) after expansion")+
  theme_bw()+
  theme_bw()+
  ggtitle(label="",subtitle="only scenarios with heritable variation in *d<sub>max</sub>*")+
  theme(legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(angle = 0),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.subtitle = element_markdown()
  )
```