---
title: 'main analysis code for "Individual variation in dispersal, and its sources, shape the fate of pushed vs. pulled range expansions"'
author: "Maxime Dahirel, Chloé Guicharnaud, Elodie Vercken (code by Maxime Dahirel)"
date:
output:
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

```{r load-packages}
library(ordinal) # CRAN v2019.12-10
library(arm) # CRAN v1.11-2
library(matrixStats) # CRAN v0.60.0

library(mgcv) # CRAN v1.8-35
library(gratia) # CRAN v0.6.0

library(MuMIn) # CRAN v1.43.17

library(tidyverse) # CRAN v1.3.1

library(cowplot) # [github::wilkelab/cowplot] v1.1.1
library(patchwork) # CRAN v1.1.1

library(ggdist) # CRAN v3.0.0

library(ggtext) # CRAN v0.1.1

library(here) # CRAN v1.0.1
```

# Introduction

After running the code in `R/generate_simulations`, we obtained, and saved, a `csv` file containing patch-level data for every simulation we ran. We can now load it, but we're going to immediately make a quick change: in the IBM, phenotypic variance input variables denote, well, the actual phenotypic variances we wanted to input. But since we only tested two levels, variance or no variance, much easier for us to work with binary variables going forward:

```{r import-data}
data <- read_csv(here("NetLogo_output", "simulation_output.csv")) %>%
  mutate(
    is.VP_dmax = as.numeric(VP_logit_dmax > 0),
    is.VP_DDD = as.numeric(VP_slope > 0)
    # if VP_slope >0, then VP_midpoint is also >0 in our experimental design
    # so only one binary variable needed for density-dependent dispersal variables
  )
```

OK, the dataset we just imported is structured so that one row = one patch in one landscape at one observed time. Only patches that were populated at least once are included (see the end of the `generate_simulations` file). The dataset contains the following variables:

- `ticks`: time step at which the observations were made (= number of generations since start)

Landscape-level variables, all constant across time (either because they are truly constant, or because they only describe initial conditions at $t=0$ and are not updated after):
              
- `start_dmax`, `start_slope`, `start_midpoint`: mean dispersal trait values at the start of the experiment (`slope` corresponds to $\alpha$ in the manuscript, `midpoint` to $\beta$)
- `VP_logit_dmax`, `VP_slope`, `VP_midpoint`: **initial** phenotypic variances
- `is.VP_dmax`, `is.VP_DDD`: binary variables describing whether `VP_logit_dmax` on one hand, and `VP_slope` and `VP_midpoint` on the other, are >0
- `heritability`: **initial** heritability (same value for all traits)
- `dispersal_mortality`: probability of dying during dispersal
- `fecundity`: hypothetical average fecundity at $N=0$ ($\exp(r_{0})$)               
- `K`: patch carrying capacity (same for all patches)
- `reproduction`: whether the simulated species was clonal or sexual (here all runs are clonal)
- `seedID`, `siminputrow`, `replicateID`: the first two variables, combined, give the third one: a unique identifier for each landscape/run
- `start_d0`, `start_dK`, `start_davg_0_K`, `start_slopeA_0_K`, `start_slopeA_0_avg`: mean values for $d_{0}$, $d_{K}$, $d_{avg}$,  the density-dependence metrics $\Delta_{K-0}$, $\Delta_{avg-0}$ in the starter population (see main text, **Fig. 1** for the definition of each of these variables)        
- `start_uncond_0_K`: same as above, for the unconditionality metric (see **Supplementary Material S6** for the definition of that metric)

- `agent`: here always "patches"                   
- `breed`: here always NA (Netlogo "breeds" are ways to have different discrete groups of individuals that play by different rules. Here there was only one breed, and data are collected at the patch level anyway)                                                  

Patch-level variables:
- `pxcor`: patch $x$ coordinate                   
- `founding`: the generation at which the patch was first populated
- `N_predispersal` and `N_postdispersal`: population size in the patch immediately before and after the dispersal phase, respectively
- `N_sedentary`: number of individuals born in the patch that stayed in it
- `N_disp_dead`: number of individuals born in the patch that died during dispersal             
- `N_allele0_pre`, `N_allele1_pre`, `N_allele0_post`, `N_allele1_post`: number of individuals with each neutral allele, pre- and post-dispersal
- `mean_genotype_logit_dmax`, `var_genotype_logit_dmax`, `mean_genotype_slope`, `var_genotype_slope`, `mean_genotype_midpoint`,  `var_genotype_midpoint`: mean and variances  of the genetic component of traits   
- `mean_dmax`, `mean_slope`, `mean_midpoint`: phenotypic means of the dispersal traits
- `mean_d0`, `mean_dK`, `mean_davg_0_K`, `mean_slopeA_0_K`, `mean_slopeA_0_avg`, `mean_uncond_0_K`: means of dispersal metrics that can be estimated from these dispersal traits (see above)     

# Data wrangling

To do our analyses, we're going to build four "processed" tables from this "raw" dataset, which we'll then merge into a final table.

## sub-table 1: expansion speed

Here we build a table that contains, for each time x replicate combination, summary information about the location of the range edge after dispersal (farthest populated patch): 

```{r tab-front}
tab_speed <- data %>%
  filter(N_postdispersal > 0) %>%
  group_by(replicateID, ticks) %>% # for each time x replicate
  summarise(edge_pxcor = max(pxcor))
```

## sub-table 2: trait values
Here we have mean traits in the range front (defined here as patches < 5 patches from the edge; see Supplementary S2 for justification)

```{r tab-trait}
tab_intermediate <- data %>%
  filter(N_predispersal > 0) %>%
  # Here we use predispersal population size since that's when traits are recorded,
  # but results are similar if one  use post-dispersal populations instead
  group_by(replicateID, ticks) %>% # for each time x replicate
  mutate(limit_pxcor = max(pxcor)) %>%
  ungroup() %>%
  filter(pxcor > limit_pxcor - 5) # and we only keep patches within the front


## from here we summarise every key phenotypic trait with weighted averages

## we first isolate variables that won't be modified by wrangling for a while
## to reduce memory size of the big table that will be manipulated
## we'll put back everything together later
tab_annotation <- tab_intermediate %>%
  select(
    replicateID,
    dispersal_mortality, heritability,
    is.VP_dmax, is.VP_DDD,
    fecundity
  ) %>%
  distinct()

## code may slow down here for a while, probably due to the size of the summarise call

tab_traits <- tab_intermediate %>%
  select(
    replicateID, ticks,
    mean_dmax, mean_davg_0_K, mean_d0,
    mean_slopeA_0_avg, mean_slopeA_0_K, mean_uncond_0_K,
    N_sedentary, N_predispersal
  ) %>%
  group_by(replicateID, ticks) %>%
  summarise(
    mean_dmax = weighted.mean(mean_dmax, w = N_predispersal, na.rm = TRUE),
    mean_davg_0_K = weighted.mean(mean_davg_0_K, w = N_predispersal, na.rm = TRUE),
    mean_d0 = weighted.mean(mean_d0, w = N_predispersal, na.rm = TRUE),
    mean_avgslope = weighted.mean(mean_slopeA_0_avg, w = N_predispersal, na.rm = TRUE),
    mean_Kslope = weighted.mean(mean_slopeA_0_K, w = N_predispersal, na.rm = TRUE),
    mean_uncond = weighted.mean(mean_uncond_0_K, w = N_predispersal, na.rm = TRUE),
    ## and the other non-trait metrics using means or sum as relevant
    N_sedentary = sum(N_sedentary),
    N_predispersal = sum(N_predispersal)
  ) %>%
  ungroup()


## finally for our analysis, we need these metrics to be in the wide format
## one column per time sampled
## needed in order to see which "moment" predicts best the observed dynamics

tab_traits <- tab_traits %>%
  ungroup() %>%
  left_join(tab_speed) %>% ## we merge back here the subtable 1 already
  pivot_wider(
    id_cols = c(
      replicateID
    ),
    names_from = ticks,
    values_from = c(
      mean_dmax, mean_davg_0_K, mean_d0,
      mean_avgslope, mean_Kslope, mean_uncond,
      edge_pxcor, N_sedentary, N_predispersal
    )
  ) %>%
  left_join(tab_annotation) # get these info variables back in
```

## sub-table 3: starting conditions

We also extract separately the values of key starting variables:
```{r starting-conditions}
tab_start <- data %>%
  select(
    start_dmax,
    start_davg_0_K,
    start_d0,
    start_avgslope = start_slopeA_0_avg,
    start_Kslope = start_slopeA_0_K,
    start_uncond = start_uncond_0_K,
    replicateID
  ) %>%
  distinct() # avoid duplicates
```

## sub-table 4: information about genetic diversity loss

Finally, we run another extraction to get information on genetic diversity decay, using the time to fixation as a metric (i.e. earliest observation of front (defined as above) only containing one of the two neutral alleles):

```{r tab-genetic}

tab_genet <- tab_intermediate %>%
  group_by(
    replicateID, ticks,
    dispersal_mortality, heritability,
    is.VP_dmax, is.VP_DDD,
    fecundity
  ) %>%
  # for each time x replicate combination, we selected only front patches
  # (see above how tab_intermediate was built)
  # and now sum the individuals with each neutral allele within the front
  summarise(
    N_allele0 = sum(N_allele0_pre),
    N_allele1 = sum(N_allele1_pre)
  ) %>%
  ungroup() %>%
  # for each time x replicate combination, genetic diversity is lost if
  # one of the two alleles is missing
  mutate(front_fixed = N_allele0 == 0 | N_allele1 == 0) %>%
  mutate(
    time_to_fix = case_when(
      front_fixed == 1 ~ ticks,
      T ~ 121 ## 121 placeholder for "not fixed yet at end of experiment"
    )
  ) %>%
  group_by(
    replicateID, dispersal_mortality,
    heritability,
    is.VP_dmax, is.VP_DDD,
    fecundity
  ) %>%
  # for each replicate, the time to fixation is
  # the earliest time no diversity was seen:
  summarise(time_to_fix = min(time_to_fix)) %>%
  ungroup() %>%
  # finally (and importantly!!) we convert it into an ordered factor:
  mutate(fixinterval = ordered(time_to_fix,
    levels = sort(unique(time_to_fix))
  ))
```

## merging the sub-tables

Now we can mix and combine these datasets for analyses (and export the merged table, since it's going to be useful for supplementaries):

```{r merge-data-tables}

tab_all <- tab_traits %>% ## tab_traits already merged with tab_speed, see above
  left_join(tab_start) %>%
  left_join(tab_genet) %>%
  mutate(mortality = factor(dispersal_mortality)) %>%
  mutate(advance = edge_pxcor_120 - edge_pxcor_100)

write_csv(tab_all, here("R_output", "processed_data_all_fecundities.csv"))
# export to be usable by supplementary.md
```

IMPORTANT! if you read the `processed_data_all_fecundities.csv` file back into R, the **first** thing to do is tell R `mortality` is a factor and `fixinterval` is an ordered factor, or else it will read them as numeric variables and this will break all sorts of stuff (see how this is done in `supplementary.rmd`)

# Figures and analyses

For the main text, we focus on the high fecundity case (the same analyses with low fecundity are in supplementary):
```{r keep-only-high-fecundity}
tab <- tab_all %>%
  filter(fecundity == 5)
```

But before analyzing these simulated data,let's draw our **Figure 1**:

## Figure 1: overview of the dispersal function

The aim of *Fig. 1*, which is in Methods, is 
- to summarise visually how our density-dispersal function behaves, and the role of each of the three parameters
- to show visually what our 2 metrics of density-dependence represent

```{r fig1-top}
## we create a table that contains variation for all 3 parameters:
p1_tab <- tibble(x = c(0:120) / 100) %>%
  expand_grid(
    dmax = c(0.2, 0.5, 0.7),
    a = c(-10, 0, 10),
    b = c(0.2, 0.5, 1.1)
  ) %>%
  mutate(group = paste(dmax, a, b)) %>%
  mutate(d = dmax / (1 + exp(-a * (x - b))))


## we make a subplot where we show what happens when dmax varies...
p1_1_dmax <- p1_tab %>%
  filter(b == 0.5 & a == 10) %>%
  ggplot() +
  geom_line(aes(x, d, group = group)) +
  geom_textbox(
    data = tibble(dmax = unique(p1_tab$dmax)),
    aes(
      x = 1,
      y = dmax - 0.05,
      label = paste("*d<sub>max</sub>* = ", dmax)
    ),
    width = grid::unit(0.3, "npc")
  ) +
  scale_x_continuous("N/K") +
  scale_y_continuous("Dispersal probability") +
  theme_bw()

## when alpha (slope) varies...
p1_2_alpha <- p1_tab %>%
  filter(b == 0.5 & dmax == 0.7) %>%
  ggplot() +
  geom_line(aes(x, d, group = group)) +
  geom_textbox(
    data = tibble(
      b = 0.5,
      dmax = 0.7,
      a = unique(p1_tab$a)
    ),
    aes(
      x = 1,
      y = dmax / (1 + exp(-a * (1 - b))),
      label = paste("*&alpha;* =", a)
    ),
    width = grid::unit(0.22, "npc")
  ) +
  scale_x_continuous("N/K") +
  scale_y_continuous("") +
  theme_bw()

### and when beta (midpoint) varies.
p1_3_beta <- p1_tab %>%
  filter(a == 10 & dmax == 0.7) %>%
  ggplot() +
  geom_line(aes(x, d, group = group)) +
  geom_textbox(
    data = tibble(
      b = unique(p1_tab$b),
      dmax = 0.7,
      a = 10
    ),
    aes(
      x = b,
      y = 0.4,
      label = paste("*&beta;* =", b)
    ),
    width = grid::unit(0.15, "npc")
  ) +
  scale_x_continuous("N/K") +
  scale_y_continuous("") +
  theme_bw()
```

For the bottom part, we're gonna need some colours. So let's make a color palette, it's going to be useful later. We're using ColorBrewer () to build a 3 colour palette that is hopefully colour friendly:

```{r colour-palette}
mypalette <- c("#1f78b4", "#b2df8a")
```

```{r fig1-bottom}
### Below that, we are going to plot two examples of functions
### and how we sample them to estimate our density-dependency metrics:

p1_tab2 <- tibble(
  dmax = c(0.734, 0.8),
  a = c(20, 6),
  b = c(0.2, 0.6)
) %>%
  expand_grid(x = rep(c(0:10) / 10, 2)) %>%
  mutate(d = dmax / (1 + exp(-a * (x - b))))


p1_4 <- p1_tab2 %>%
  ggplot(aes(x = x, y = d)) +
  geom_line(aes(col = factor(dmax))) +
  geom_point(aes(col = factor(dmax), shape = factor(dmax)), size = 5, fill = "white") +
  scale_x_continuous("N/K") +
  scale_y_continuous("Dispersal probability") +
  scale_colour_manual(values = mypalette) +
  scale_shape_discrete() +
  theme_bw()

# you can check colors using colorblindr::cvd_grid(p1_4) if you have that package installed

p1_5 <- p1_tab2 %>%
  group_by(dmax) %>%
  summarise(
    d0 = subset(d, x == 0),
    dK = subset(d, x == 1),
    davg = mean(d)
  ) %>%
  pivot_longer(cols = c(d0, dK, davg)) %>%
  mutate(name = fct_recode(name,
    `*d<sub>0</sub>*` = "d0",
    `*d<sub>avg</sub>*` = "davg",
    `*d<sub>K</sub>*` = "dK"
  )) %>%
  ggplot() +
  geom_point(aes(name, value, col = factor(dmax), shape = factor(dmax)),
    size = 5, fill = "white", position = position_dodge(width = 0.4)
  ) +
  scale_x_discrete("dispersal variable") +
  scale_y_continuous("value") +
  scale_colour_manual(values = mypalette) +
  scale_shape_discrete() +
  theme_bw() +
  theme(axis.text.x = element_markdown())
```


```{r fig1}
(p1_1_dmax | p1_2_alpha | p1_3_beta) / (p1_4 + p1_5) &
  theme(legend.position = "none") &
  plot_annotation(tag_levels = "A")
```


## analyses - models

### velocity: fitting GAMs to simulated data

As mentioned in the main text, we run seven models to try to predict velocity between $t=100$ and $t=120$ (Why these times? because we consider velocities have reached equilibrium by then. Why do we think that? See Supplementary S3). Three models assumed that traits post-evolution (when there is evolution) matter, three assume traits before evolution matter, and a final one acts as a "null" of sort, by assuming traits have no impact, only mortality.

Because patches are discrete, and individuals advance at most 1 patch per generation, we assume a binomial model:

```{r velocity-models}

## the models assuming traits post-evolution matter
mod_100 <- gam(cbind(advance, 20 - advance) ~
s(mean_d0_100, by = mortality) +
  s(mean_avgslope_100, by = mortality) +
  mortality,
family = binomial, data = tab, method = "REML"
# we follow e.g. Pedersen et al 10.7717/peerj.6876 in using REML for GAMs
)

mod_100_K <- gam(cbind(advance, 20 - advance) ~
s(mean_d0_100, by = mortality) +
  s(mean_Kslope_100, by = mortality) +
  mortality,
family = binomial, data = tab, method = "REML"
)

mod_100_noDDD <- gam(cbind(advance, 20 - advance) ~
s(mean_d0_100, by = mortality) +
  mortality,
family = binomial, data = tab, method = "REML"
)

## the models assuming initial traits matter
mod_00 <- gam(cbind(advance, 20 - advance) ~
s(start_d0, by = mortality) +
  s(start_avgslope, by = mortality) +
  mortality,
family = binomial, data = tab, method = "REML"
)

mod_00_K <- gam(cbind(advance, 20 - advance) ~
s(start_d0, by = mortality) +
  s(start_Kslope, by = mortality) +
  mortality,
family = binomial, data = tab, method = "REML"
)

mod_00_noDDD <- gam(cbind(advance, 20 - advance) ~
s(start_d0, by = mortality) +
  mortality,
family = binomial, data = tab, method = "REML"
)

## the mortality only model
mod_null <- gam(cbind(advance, 20 - advance) ~
mortality,
family = binomial, data = tab, method = "REML"
)
```

### a minor detour: underdispersion?

If you look at the model diagnostics, you'll see data look a bit underdispersed, compared to the model. Yes, *under*dispersed,

```{r dispersion-test}
## function is copied from Ben Bolker FAQ
## https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#overdispersion
## with only one modification to test for under instead of overdispersion
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = TRUE) # lower.tail = TRUE instead of FALSE is the only change
  c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
}

overdisp_fun(mod_100)
```

See also comment from Bob Carpenter here: https://discourse.mc-stan.org/t/underdispersed-binomial-glm/1540

The general idea is that underdispersion relative to the binomial may be the sign that the trials that are collapsed together for each response (the generations, here) are not independent. This makes sense here: probability of advancing depends on densities at the front, and we can expect these to be temporally autocorrelated. Intuitively this should not be as big a problem as overdispersion: underdispersion means the model will exaggerate variance compared to data, which should lead to more conservative results?

But to be sure, let's use an ordinal model as a quick comparison: 
(see e.g. https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#underdispersion)

```{r ordinal-velocity}
tab_ordinal <- tab %>%
  mutate(
    scale_d0_100 = scale(mean_d0_100)[, 1],
    scale_d0_start = scale(start_d0)[, 1],
    s_avgslope_100 = mean_avgslope_100 / sd(mean_avgslope_100),
    s_avgslope_start = start_avgslope / sd(start_avgslope),
    s_Kslope_100 = mean_Kslope_100 / sd(mean_Kslope_100),
    s_Kslope_start = start_Kslope / sd(start_Kslope)
  )

mod_100_ordinal <- clm(ordered(advance) ~
(scale_d0_100 + s_avgslope_100) * mortality,
link = "cloglog", data = tab_ordinal
)

summary(mod_100_ordinal)
draw(mod_100)
```

If you compare the summary of the ordinal model with the effect plot of the binomial GAM, the conclusions are pretty similar indeed (you can also check with the other models). But since we can't use splines with the ordinal model, let's keep the binomial models as our main ones.

### velocity: model selection among GAMs

To find out which of our seven GAMs is the best, let's make a model selection table. We first create a function to extract % of deviance explained and R^2^ from the model objects, to add them to the selection table:

```{r GAM-R2}
gam_dev.expl <- function(mod) {
  return(summary(mod)$dev.expl)
}

gam_r2 <- function(mod) {
  return(summary(mod)$r.sq)
}
```

Then we can make the model selection table:

```{r velocity-modsel}
modsel_velocity <- model.sel(list(
  mod_100 = mod_100,
  mod_100_K = mod_100_K,
  mod_100_noDDD = mod_100_noDDD,
  mod_00 = mod_00,
  mod_00_K = mod_00_K,
  mod_00_noDDD = mod_00_noDDD,
  mod_null = mod_null
), extra = list(gam_dev.expl, gam_r2)) %>%
  # MuMIn::model.sel puts model names as a row name rather than as a variable
  # let's correct this:
  rownames_to_column(var = "model_name") %>%
  as_tibble() %>%
  # then let's add variables that tell us about what's in each model
  # re: does it include density-dependency? use traits pre/post evolution?
  mutate(
    is.evo = case_when(
      str_detect(model_name, "null") ~ NA_character_,
      str_detect(model_name, "_100") ~ "after evolution (t = 100)",
      T ~ "before evolution (t = 0)"
    ),
    is.DDD = case_when(
      str_detect(model_name, "null") ~ NA_character_,
      str_detect(model_name, "_noDDD") ~ "ignored",
      str_detect(model_name, "_K") ~ "*&Delta;<sub>K-0</sub>* (shape-indep.)",
      T ~ "*&Delta;<sub>avg-0</sub>* (shape-dep.)"
    )
  ) %>%
  mutate(is.evo = fct_relevel(is.evo, "before evolution (t = 0)", after = 0)) %>%
  mutate(is.DDD = fct_relevel(is.DDD, "ignored", after = 0)) %>%
  select(model_name, gam_dev.expl:is.DDD) # let's remove all the coef columns
# that model.sel adds automatically

modsel_velocity
```

It is extremely clear here what model is the best, independently of the criterion used (it is "density-dependence included, shape-dependent, post-evolution"). We keep that in mind for later, and move to the analysis of genetic diversity data.

### genetic diversity: fitting ordinal models

Our measure of the capacity of expansions to keep **neutral** genetic diversity at the front is... the time they take to lose it, i.e. having only one allele at the front. This is time to event data, so it should be easy to analyse using survival models. Except we only sampled expansions every 20 generations to save memory. But no worry, this is where discrete time-to-event models come to the rescue. A simple way to fit these is to fit an ordinal model, especially with a cloglog link, since it becomes basically a discrete proportional hazards model (see e.g. Bürkner and Vuorre 2019 doi: 10.1177/2515245918823199). So let's do that, with seven model structures, just like velocity:

```{r genetic-models}
mod_div_00 <- clm(fixinterval ~
mortality * (start_d0 + start_avgslope),
data = filter(tab, edge_pxcor_120 > 0), # one front did not advance
link = "cloglog"
)

mod_div_00 <- clm(fixinterval ~
mortality + mortality:(start_d0 + start_avgslope),
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)

mod_div_00_K <- clm(fixinterval ~
mortality + mortality:(start_d0 + start_Kslope),
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)

mod_div_00_noDDD <- clm(fixinterval ~
mortality + mortality:(start_d0),
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)

mod_div_100 <- clm(fixinterval ~
mortality + mortality:(mean_d0_100 + mean_avgslope_100),
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)

mod_div_100_K <- clm(fixinterval ~
mortality + mortality:(mean_d0_100 + mean_Kslope_100),
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)

mod_div_100_noDDD <- clm(fixinterval ~
mortality + mortality:(mean_d0_100),
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)

mod_div_null <- clm(fixinterval ~
mortality,
data = filter(tab, edge_pxcor_120 > 0),
link = "cloglog"
)
```

### neutral genetic diversity: model selection table

Again, let's compare the models. And again, in addition to AIC, let's build a function to add a R^2^ to the model selection table. Except that, since it's ordinal, it's going to be a pseudo R^2^ here: based on the regression of observed categories converted to numbers and predictions:

```{r ordinal-R2}

clm_pseudo_r2 <- function(model, data = tab) { ## given a model and the *original* data
  # we make predictions
  pp <- predict(model, newdata = data %>% select(-fixinterval))$fit
  # it gets us probabilities for each categories, we sum them to get the rank of the average category
  preds <- 1 * pp[, 1] + 2 * pp[, 2] + 3 * pp[, 3] + 4 * pp[, 4] + pp[, 5] + 6 * pp[, 6] + 7 * pp[, 7]
  obs <- as.numeric(data$fixinterval)
  # then we use that and the rank of the observed category in a linear regression
  return(summary(lm(obs ~ preds))$r.squared)
}
```

Then we build the model table pretty much the same way as for velocity:

```{r ordinal-modsel}
modsel_diversity <- model.sel(list(
  mod_div_100 = mod_div_100,
  mod_div_100_K = mod_div_100_K,
  mod_div_100_noDDD = mod_div_100_noDDD,
  mod_div_00 = mod_div_00,
  mod_div_00_K = mod_div_00_K,
  mod_div_00_noDDD = mod_div_00_noDDD,
  mod_div_null = mod_div_null
), extra = list(clm_pseudo_r2)) %>%
  rownames_to_column(var = "model_name") %>%
  as_tibble() %>%
  mutate(
    is.evo = case_when(
      str_detect(model_name, "null") ~ NA_character_,
      str_detect(model_name, "_100") ~ "after evolution (t = 100)",
      T ~ "before evolution (t = 0)"
    ),
    is.DDD = case_when(
      str_detect(model_name, "null") ~ NA_character_,
      str_detect(model_name, "_noDDD") ~ "ignored",
      str_detect(model_name, "_K") ~ "*&Delta;<sub>K-0</sub>* (shape-indep.)",
      T ~ "*&Delta;<sub>avg-0</sub>* (shape-dep.)"
    )
  ) %>%
  mutate(is.evo = fct_relevel(is.evo, "before evolution (t = 0)", after = 0)) %>%
  mutate(is.DDD = fct_relevel(is.DDD, "ignored", after = 0)) %>%
  select(model_name, clm_pseudo_r2:is.DDD) # let's remove all the coef columns
# that model.sel adds automatically

modsel_diversity
```

### Figure 2 : model selection

Now that we have model selection tables for both responses, we can build a figure summing up how each model structure performs for each:

```{r fig2-palette}
mypalette <- c("#1b9e77", "#d95f02", "#7570b3")
```


```{r fig2}

p2_1 <- modsel_velocity %>%
  filter(model_name != "mod_null") %>%
  ggplot() +
  geom_point(aes(is.evo, gam_r2,
    col = is.DDD,
    shape = is.DDD
  ),
  size = 4
  ) +
  geom_line(aes(is.evo, gam_r2,
    group = is.DDD,
    col = is.DDD
  )) +
  geom_hline(yintercept = summary(mod_null)$r.sq, lty = 2) + ## the aicc of the mortality only model
  scale_y_continuous("R^2") +
  scale_colour_manual(values = mypalette) +
  scale_shape_discrete() +
  labs(title = "expansion velocity")


p2_2 <- modsel_diversity %>%
  filter(model_name != "mod_div_null") %>%
  ggplot() +
  geom_point(aes(is.evo, clm_pseudo_r2,
    col = is.DDD,
    shape = is.DDD
  ),
  size = 4
  ) +
  geom_line(aes(is.evo, clm_pseudo_r2,
    group = is.DDD,
    col = is.DDD
  )) +
  geom_hline(yintercept = clm_pseudo_r2(model = mod_div_null), lty = 2) + ## the aicc of the mortality only model
  scale_y_continuous("pseudo-R^2") +
  scale_colour_manual(values = mypalette) +
  scale_shape_discrete() +
  labs(title = "neutral diversity")

(p2_1 + p2_2 + plot_layout(guides = "collect")) & scale_x_discrete("when are trait covariables sampled?") &
  labs(col = "Density dependence?", shape = "Density dependence?") & theme_bw() &
  theme(
    axis.title.y = element_markdown(),
    legend.text = element_markdown()
  )
```

## Figure 3: effect of d0 and density-dependence on velocities

For Figure 3, we want to use the best velocity model to display how traits influence expected speed. We do it in two steps, one for the effect of d0, one for the effect of density-dependence. For each step, we build a grid with the range of possible values for the focal predictor trait, fix the other to a meaningful value (0 for the density-dependence when predicting along d0, the average for d0 when predicting along density-dependence), and make predictions. We then display them:

```{r fig3}

# we create a grid with the full range of d0 and mortalities
# but assuming no density dependence (avgslope = 0)
newdata <- tibble(
  mean_avgslope_100 = 0,
  mean_d0_100 = c(0:100) / 100
) %>%
  expand_grid(mortality = c("0.1", "0.5", "0.9")) %>%
  mutate(row = 1:length(mortality))

# and use it to predict expansion speed and plot it
p1 <- fitted_samples(mod_100, n = 1000, newdata = newdata, seed = 42, scale = "response") %>%
  left_join(newdata) %>%
  ggplot() +
  stat_lineribbon(aes(mean_d0_100, fitted), .width = c(0.001, 0.95), fill = "grey") +
  scale_y_continuous("predicted expansion velocity (patches/generation)") +
  scale_x_continuous("mean *d<sub>0</sub>* at *t* = 100") +
  facet_grid(
    rows = vars(paste("*m* = ", mortality)),
    cols = vars("effect of *d<sub>0</sub>*")
  ) +
  coord_cartesian(ylim = c(0, 1))

# then we do the same, this time let avgslope vary, and fixing d0 to its average
newdata <- tibble(
  mean_avgslope_100 = seq(
    from = range(tab$mean_avgslope_100)[1],
    to = range(tab$mean_avgslope_100)[2],
    length.out = 100
  ),
  mean_d0_100 = mean(tab$mean_d0_100)
) %>%
  expand_grid(mortality = c("0.1", "0.5", "0.9")) %>%
  mutate(row = 1:length(mortality))

# then we predict expansion speeds again and plot them again
p2 <- fitted_samples(mod_100, n = 1000, newdata = newdata, seed = 42, scale = "response") %>%
  left_join(newdata) %>%
  ggplot() +
  stat_lineribbon(aes(mean_avgslope_100, fitted), .width = c(0.001, 0.95), fill = "grey") +
  scale_y_continuous("") +
  scale_x_continuous("mean *&Delta;<sub>avg-0</sub>* at *t* = 100") +
  facet_grid(
    rows = vars(paste("*m* = ", mortality)),
    cols = vars("effect of density-dependence *&Delta;<sub>avg-0</sub>*")
  ) +
  coord_cartesian(ylim = c(0, 1))

(p1 | p2) &
  theme_bw() &
  theme(
    axis.title.x = element_markdown(),
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown()
  )
```


## Figure 4: effect of d0 and density-dependence on neutral diversity

In principle, we do basically the same as in Figure 3, except for genetic diversity: create new grids, predict for these grids based on the best model, plot predictions. However, because these are ordinal models, predictions look a bit different: we get one probability for each ordinal category. We could average to get the average category/time to fixation, but this does not really make sense since the final category is an open-ended one "time to fixation > 120".

So instead, we display the full distribution of the predictions:

```{r fig4}
newdata <- tibble(
  start_avgslope = 0,
  start_d0 = c(0:50) / 100
) %>%
  expand_grid(mortality = c("0.1", "0.5", "0.9")) %>%
  mutate(row = 1:length(mortality))

p1 <- cbind(newdata, predict(mod_div_00, newdata = newdata)$fit) %>%
  pivot_longer(cols = c(`20`, `40`, `60`, `80`, `100`, `120`, `121`)) %>%
  mutate(name = fct_recode(name,
    `< 20` = "20",
    `21 to 40` = "40",
    `41 to 60` = "60",
    `61 to 80` = "80",
    `81 to 100` = "100",
    `101 to 120` = "120",
    `> 120` = "121"
  )) %>%
  mutate(name = str_replace(as.character(name), "< 20", "\u2264 20")) %>%
  mutate(name = fct_relevel(
    name,
    "\u2264 20",
    "21 to 40",
    "41 to 60",
    "61 to 80",
    "81 to 100",
    "101 to 120",
    "> 120"
  )) %>%
  ggplot() +
  geom_area(aes(x = start_d0, y = value, fill = name), position = "fill") +
  scale_y_continuous("") +
  scale_x_continuous("mean *d<sub>0</sub>* at *t* = 0") +
  guides(fill = "none") +
  facet_grid(
    rows = vars(paste("*m* = ", mortality)),
    cols = vars("effect of *d<sub>0</sub>*")
  ) +
  coord_cartesian(ylim = c(0, 1))

newdata <- tibble(
  start_avgslope = seq(
    from = range(tab$start_avgslope)[1],
    to = range(tab$start_avgslope)[2],
    length.out = 100
  ),
  start_d0 = mean(tab$start_d0)
) %>%
  expand_grid(mortality = c("0.1", "0.5", "0.9")) %>%
  mutate(row = 1:length(mortality))

p2 <- cbind(newdata, predict(mod_div_00, newdata = newdata)$fit) %>%
  pivot_longer(cols = c(`20`, `40`, `60`, `80`, `100`, `120`, `121`)) %>%
  mutate(name = fct_recode(name,
    `< 20` = "20",
    `21 to 40` = "40",
    `41 to 60` = "60",
    `61 to 80` = "80",
    `81 to 100` = "100",
    `101 to 120` = "120",
    `> 120` = "121"
  )) %>%
  mutate(name = str_replace(as.character(name), "< 20", "\u2264 20")) %>%
  mutate(name = fct_relevel(
    name,
    "\u2264 20",
    "21 to 40",
    "41 to 60",
    "61 to 80",
    "81 to 100",
    "101 to 120",
    "> 120"
  )) %>%
  ggplot() +
  geom_area(aes(x = start_avgslope, y = value, fill = name), position = "fill") +
  scale_y_continuous("") +
  scale_x_continuous("mean *&Delta;<sub>avg-0</sub>* at *t* = 0") +
  labs(fill = "generations before allele fixation:") +
  facet_grid(
    rows = vars(paste("*m* = ", mortality)),
    cols = vars("effect of density-dependence *&Delta;<sub>avg-0</sub>*")
  ) +
  coord_cartesian(ylim = c(0, 1))



(p1 | p2) &
  theme_bw() &
  scale_fill_brewer(palette = "Greys") &
  plot_layout(guides = "collect") &
  theme(
    axis.title.x = element_markdown(),
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown()
  )
```

## Figure 5: evolution of dmax

Here we draw a plot where we show the t=100 value of dispersal capacity ($d_{max}$), depending on heritability, mortality and source of variation. We plot the replicate values in grey, and the overall averages in black. We also add the mean starting value (0.5 for everyone) as a dashed line, as a reminder.

Note that for this plot, we obviously excluded replicates where the initial phenotypic variance for $d_{max}$ was 0.


```{r fig5}
tab %>%
  filter(is.VP_dmax == 1) %>% # does not make sense to include replicates with no variation in dmax
  mutate(is.VP_DDD = factor(is.VP_DDD)) %>%
  ggplot() +
  geom_hline(yintercept = 0.5, lty = 2) +
  geom_jitter(aes(is.VP_DDD, mean_dmax_100),
    col = "grey80"
  ) +
  stat_summary(aes(is.VP_DDD, mean_dmax_100),
    fun = "mean", fun.min = "mean", fun.max = "mean", size = 1
  ) +
  scale_x_discrete("individual variation in dispersal traits",
    labels = c(
      "in *d<sub>max</sub>* only",
      "in *&alpha;*, *&beta;*, and *d<sub>max</sub>*"
    )
  ) +
  scale_y_continuous("mean *d<sub>max</sub>* at the front at *t* = 100") +
  facet_grid(
    rows = vars(paste("*m* =", dispersal_mortality)),
    cols = vars(paste("*h^2* =", heritability))
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(),
    axis.text.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.title = element_markdown()
  )
```

## Figure 6: evolution of density-dependence

Here we do a broadly similar thing as in Figure 5, but for our density-dependent metric $\Delta_{avg-0}$. So there are two big changes actually:
- first, the initial value of the metric varies here depending on replicates, so the base plot is different. We use a (z-x)~x scatterplot where z=trait after evolution (t=100), x=trait at t0. Deviation from the horizontal line denoting no change indicates evolution. In addition, we fit the corresponding linear regression on each subplot

- second, there are three ways phenotype can vary and influence density-dependence ($d_{max}$ only, matching traits only, both), instead of 2 for $d_{max}$ ($d_{max}$ only, $d_{max}$ and matching traits). So the plot is correspondingly bigger.

Finally, it is possible to identify regions of plotted space that are impossible to evolve into, because a density-dependence that strong would imply dmax higher than the maximal possible dmax for the replicate. For instance, when $d_{max}$ cannot evolve, the range of possible dispersal rates is [0;0.5], so the range of possible slopes $\Delta_{avg-0}$ is [-0.5;0.5]. If the initial slope is, say, 0.2, then increases in slopes higher than +0.3 are not possible, for instance. So we shaded these "impossible areas" in grey, to give a better idea of the space available to evolution.


```{r fig6}
p11 <- tab %>%
  filter(is.VP_dmax == 1 & is.VP_DDD == 1) %>%
  ggplot() +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  # denoting the strategies that are possible (davg>dmax not possible):
  geom_ribbon( # shading impossible region
    data = tibble(x = c(-100:100) / 100),
    aes(x = x, ymin = 1 - x, ymax = 1),
    fill = "grey80", col = "black"
  ) +
  geom_ribbon( # cont'd
    data = tibble(x = c(-100:100) / 100),
    aes(x = x, ymin = -1 - x, ymax = -1),
    fill = "grey80", col = "black"
  ) +
  # plotting replicates
  geom_point(aes(
    start_avgslope,
    mean_avgslope_100 - start_avgslope
  ),
  col = "grey80"
  ) +
  # plotting regression
  geom_smooth(aes(
    start_avgslope,
    mean_avgslope_100 - start_avgslope
  ),
  method = "lm",
  col = "black", fill = "#d95f02"
  ) +
  labs(title = "C: Individual variation in *&alpha;*, *&beta;*, and *d<sub>max</sub>*") +
  scale_y_continuous("") +
  scale_x_continuous("Mean density-dependence (*&Delta;<sub>avg-0</sub>*) at *t* = 0") +
  facet_grid(
    rows = vars(paste("*m* =", dispersal_mortality)),
    cols = vars(paste("*h^2* =", heritability))
  ) +
  coord_cartesian(xlim = c(-0.3, 0.3), ylim = c(-0.25, 0.8)) +
  theme_bw()

p10 <- tab %>%
  filter(is.VP_dmax == 1 & is.VP_DDD == 0) %>%
  ggplot() +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_ribbon(
    data = tibble(x = c(-100:100) / 100),
    aes(x = x, ymin = 1 - x, ymax = 1),
    fill = "grey80", col = "black"
  ) +
  geom_ribbon(
    data = tibble(x = c(-100:100) / 100),
    aes(x = x, ymin = -1 - x, ymax = -1),
    fill = "grey80", col = "black"
  ) +
  # denotes the strategies possibles (davg>dmax not possible)
  geom_point(aes(
    start_avgslope,
    mean_avgslope_100 - start_avgslope
  ),
  col = "grey80"
  ) +
  geom_smooth(aes(
    start_avgslope,
    mean_avgslope_100 - start_avgslope
  ),
  method = "lm",
  col = "black", fill = "#d95f02"
  ) +
  labs(title = "B: Individual variation in *d<sub>max</sub>*") +
  scale_x_continuous("Mean density-dependence (*&Delta;<sub>avg-0</sub>*) at *t* = 0") +
  scale_y_continuous("Change in mean *&Delta;<sub>avg-0</sub>* at front between *t* = 0 and *t* = 100") +
  facet_grid(
    rows = vars(paste("*m* =", dispersal_mortality)),
    cols = vars(paste("*h^2* =", heritability))
  ) +
  coord_cartesian(xlim = c(-0.3, 0.3), ylim = c(-0.25, 0.8)) +
  theme_bw()

p01 <- tab %>%
  filter(is.VP_dmax == 0 & is.VP_DDD == 1) %>%
  ggplot() +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_ribbon(
    data = tibble(x = c(-100:100) / 100),
    aes(x = x, ymin = 0.5 - x, ymax = 1), # impossible space larger because dmax cannot evolve
    fill = "grey80", col = "black"
  ) +
  geom_ribbon(
    data = tibble(x = c(-100:100) / 100),
    aes(x = x, ymin = -0.5 - x, ymax = -1),
    fill = "grey80", col = "black"
  ) +
  geom_point(aes(
    start_avgslope,
    mean_avgslope_100 - start_avgslope
  ),
  col = "grey80"
  ) +
  geom_smooth(aes(
    start_avgslope,
    mean_avgslope_100 - start_avgslope
  ),
  method = "lm",
  col = "black", fill = "#d95f02"
  ) +
  scale_y_continuous("Change in mean *&Delta;<sub>avg-0</sub>* at front between *t* = 0 and *t* = 100") +
  scale_x_continuous("") +
  ggtitle("A: Individual variation in *&alpha;* and *&beta;*") +
  facet_grid(
    rows = vars(paste("*m* =", dispersal_mortality)),
    cols = vars(paste("*h^2* =", heritability))
  ) +
  coord_cartesian(xlim = c(-0.3, 0.3), ylim = c(-0.25, 0.8)) +
  theme_bw()


p_explainer <- ggplot() +
  geom_tile(aes(x = 0, y = 0, width = 2.2, height = 2), fill = "white", col = "black") +
  scale_x_continuous("", breaks = 0) +
  scale_y_continuous("", breaks = 0) +
  geom_segment(aes(x = 0, xend = 0, y = 1, yend = -1), lty = 2) +
  geom_segment(aes(x = -1.1, xend = 1.1, y = 0, yend = 0), lty = 2) +
  geom_label(aes(x = 0.55, y = -1, label = "starts \n(more) pushed"), size = 3.5) +
  geom_label(aes(x = -0.5, y = -1, label = "starts \n(more) pulled"), size = 3.5) +
  geom_label(aes(x = -1.2, y = 0.37, label = "becomes \n(more) pushed"), size = 3.5) +
  geom_label(aes(x = -1.2, y = -0.37, label = "becomes \n(more) pulled"), size = 3.5) +
  coord_cartesian(xlim = c(-1.6, 1.5), ylim = c(-1.4, 1.4)) +
  theme_void()

layout <- "
AAA###
AAABBB
AAABBB
CCCDDD
CCCDDD
CCCDDD
"

p01 + p_explainer + p10 + p11 +
  plot_layout(design = layout) &
  theme(
    strip.text.x = element_markdown(),
    strip.text.y = element_markdown(),
    axis.title.x = element_markdown(),
    axis.title.y = element_markdown(),
    plot.title = element_markdown()
  )
```
