---
title: 'simulation management code for "Individual variation in dispersal, and its sources, shape the fate of pushed vs. pulled range expansions"'
author: "Maxime Dahirel, Chloé Guicharnaud, Elodie Vercken (code by Maxime Dahirel and Chloé Guicharnaud)"
date:
output:
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

```{r load-packages}
library(nlrx) # CRAN v0.4.2
library(tidyverse) # CRAN v1.3.1
library(here) # CRAN v1.0.1
```

This is the code used to "talk" with the NetLogo simulation and make it run the experiment described in the Methods and Supplementary Materials of the Manuscript (see there and the Info pane of the NetLogo model for details). Briefly, we simulate asexual haploid individuals spreading in a linear discrete landscape. Runs differ on whether individuals very or not in traits shaping dispersal and its density-dependence, on whether that variation is heritable (and how much), and on the costs of dispersal.

# Setup

First, let's tell R where is your NetLogo version, what version number it is, for how many generations do we want the model to run, and how many replicates we want for each parameter combination:

```{r general-information}
####
NLPATH <- "C:/Program Files/NetLogo 6.2.0" # change as needed
NLVERSION <- "6.2.0" # change as needed
nreplicates <- 10
duration <- 120
```

We now need to tell R where the actual NetLogo file of the model is (The reader is strongly encouraged to go look at it, especially the Info pane):

```{r model-path}
nl_object <- nl(
  nlversion = NLVERSION, nlpath = NLPATH,
  modelpath = here("NetLogo_model", "range_expansion.nlogo"),
  jvmmem = 1024
)
```

Then, let's set the experimental design:

- `VP_logit_dmax`, `VP_slope` and `VP_midpoint` are the initial phenotypic variances for the traits underlying the dispersal function
- `heritability` is the proportion of that variance that is, at t=0, attributable to genetic factors versus noise
- `start_dmax`, `start_slope`, `start_midpoint` are the starting median values for the dispersal traits (note how dmax is entered **not** on the logit scale here _ compared to VP_ easier and more intuitive to come up with "natural" starting points this way, converted to logit internally)
- `dispersal_mortality` is the probability of dying if you disperse 
- `reproduction` governs the type of reproduction and is fixed to "clonal" here (the NetLogo code allows it to be "sexual")`
- `fecundity` is the theoretical mean *per capita* fecundity at density 0
- `K` is patch carrying capacity

Again, see the Methods, Supplementary Materials and the Info pane for more details.

```{r design-grid}
### designing the experiment
trt_grid <- tibble(
  VP_logit_dmax = c(0, 1.5^2, 1.5^2, 1.5^2, 0, 0, 0, 1.5^2, 1.5^2, 1.5^2),
  VP_slope = c(0, 6^2, 6^2, 6^2, 6^2, 6^2, 6^2, 0, 0, 0),
  VP_midpoint = c(0, 0.5^2, 0.5^2, 0.5^2, 0.5^2, 0.5^2, 0.5^2, 0, 0, 0),
  heritability = c(0, 0, 0.33, 1, 0, 0.33, 1, 0, 0.33, 1)
  # (because if V_P=0, then any h^2 will give the same V_A than h^2=0)
) %>%
  expand_grid(
    start_dmax = 0.5, # median + VP here give relatively uniform distri over ]0-1[
    start_slope = c(-6, 0, 6), # 0, and + or -1SD from 0
    ## also, |6| as slope mean we go from 0.05dmax to 0.95dmax over a range of K centred on midpoint
    start_midpoint = c(0, 0.5, 1), # same logic here
    dispersal_mortality = c(0.1, 0.5, 0.9),
    reproduction = c("clonal"),
    fecundity = c(1.5, 5),
    ## more or less arbitrary, references previous studies/models
    K = 250 ## same
  ) %>%
  mutate(
    reproduction_nl = paste('\"', reproduction, '\"', sep = "")
    # adds the quote and escape needed to export the string to Netlogo
  )
```

Now we need to add all of this (plus information on the variables that stay constant, which output variables will we save...) to the experiment slot of the NL object (we encourage readers to look at the info pane in the .nlogo model file for details):

```{r set-experiment}

nl_object@experiment <- experiment(
  expname = "experiment-h2-2020",
  outpath = here("NetLogo_output"),
  repetition = 1,
  tickmetrics = "true",
  idsetup = "setup",
  idgo = "go",
  stopcond = "not any? turtles",
  runtime = duration,
  evalticks = c(1, 20, 40, 60, 80, 100, 120),
  # the evalticks = 1 is to guarantee something is saved every time
  # even in runs that never produce any valid value for one of metrics
  # code crashes if some metrics.patches report nothing
  ### metrics and metrics. patches: what we want the model to report back
  metrics = c(
    "ticks",
    "start_d0", "start_dK", "start_davg_0_K",
    "start_slopeA_0_K", "start_slopeA_0_avg", "start_uncond_0_K"
  ),
  metrics.patches = c(
    "pxcor", "founding",
    "N_predispersal", "N_postdispersal",
    "N_sedentary", "N_disp_dead",
    "N_allele0_pre", "N_allele1_pre",
    "N_allele0_post", "N_allele1_post",
    "mean_genotype_logit_dmax", "var_genotype_logit_dmax",
    "mean_genotype_slope", "var_genotype_slope",
    "mean_genotype_midpoint", "var_genotype_midpoint",
    "mean_dmax", "mean_slope", "mean_midpoint",
    "mean_d0", "mean_dK", "mean_davg_0_K",
    "mean_slopeA_0_K", "mean_slopeA_0_avg", "mean_uncond_0_K"
  ),
  ### constants and variables: what we feed the models
  constants = list(
    "duration" = duration
  ),
  variables = list( ### needs to be a nested list
    "start_dmax" = list(values = trt_grid$start_dmax),
    "start_slope" = list(values = trt_grid$start_slope),
    "start_midpoint" = list(values = trt_grid$start_midpoint),
    "VP_logit_dmax" = list(values = trt_grid$VP_logit_dmax),
    "VP_slope" = list(values = trt_grid$VP_slope),
    "VP_midpoint" = list(values = trt_grid$VP_midpoint),
    "heritability" = list(values = trt_grid$heritability),
    "dispersal_mortality" = list(values = trt_grid$dispersal_mortality),
    "fecundity" = list(values = trt_grid$fecundity),
    "K" = list(values = trt_grid$K),
    "reproduction" = list(values = trt_grid$reproduction_nl)
  )
)
```


And we're go for running the simulations! We set a "distinct" simulation design (see `nlrx` help).
(Careful, script is not parallelized, so running the final version is expected to take up to a few hours <!-- about 20 hours on my laptop-->)

```{r running-simulations}

# set.seed(1989) ## we can set a seed here to guarantee the set of seeds selected below in nseeds is the same everytime
nl_object@simdesign <- simdesign_distinct(nl = nl_object, nseeds = nreplicates)
nl_object@simdesign@simseeds <- 1:nreplicates # or we can force rewrite the nseeds here
# (to guarantee they are all actually different)(sometimes when drawing seeds at random you will have duplicates when you don't want to)
length(unique(nl_object@simdesign@simseeds)) == nreplicates # either way, check that all generated seeds are different
exp_h2 <- run_nl_all(nl_object)

save(list = c("exp_h2"), file = here("Netlogo_output", "raw_output.Rdata"))
```

# Output

We can now process the result for the main analyses. first we start by extracting the variables we need:

```{r processing-output-presaving}
tab <- exp_h2 %>%
  select(ticks,
    start_dmax, start_slope, start_midpoint,
    VP_logit_dmax, VP_slope, VP_midpoint,
    heritability, dispersal_mortality,
    fecundity, K, reproduction,
    seedID = `random-seed`, siminputrow,
    metrics.patches,
    start_d0, start_dK, start_davg_0_K,
    start_slopeA_0_K, start_slopeA_0_avg, start_uncond_0_K
  ) %>%
  mutate(replicateID = paste(siminputrow, seedID, sep = "_")) %>%
  unnest(cols = c(metrics.patches)) %>%
  filter(founding >= 0) %>% # excluding never-populated patches
  mutate(
    #### IMPORTANT: replace the -999 (used as NA placeholders in Netlogo model) by actual NAs
    across(.cols = everything(), na_if, -999)
  )
```

then we save the file for later use

```{r saving-output}
write_csv(tab, file = here("NetLogo_output", "simulation_output.csv"))
```

We can now use the `analyse_simulations` and `supplementary` scripts to analyse the model outputs.
