---
title: 'Pushed expansions and heritability- individual-based model, model creation'
author: "Maxime Dahirel & Chloé Guicharnaud"
date:
output:
  html_document:
    theme: yeti
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

```{r load-packages}
library(nlrx)
library(arm)
library(tidyverse)
library(cowplot)
library(here)
```

# Introduction

<!--to detail-->

# Setup

First, let's tell R where is your NetLogo version, what version number it is, for how many generations do we want the model to run, and how many replicates we want for each parameter combination:

```{r general-information}
####
NLPATH <- "C:/Program Files/NetLogo 6.1.1" # change as needed
NLVERSION <- "6.1.1"                       # change as needed
nreplicates <- 1   # test run
duration <- 10     # test run
```

We now need to tell R where the actual NetLogo file of the model is (The reader is strongly encouraged to go look at it, especially the Info pane):

```{r model-path}
nl_object <- nl(
  nlversion = NLVERSION, nlpath = NLPATH,
  modelpath = here("NetLogo_model","range_expansion.nlogo"),
  jvmmem = 1024
)
```


Then, let's set the experimental design. <!--to detail-->

```{r design-grid}
### designing the experiment
trt_grid <- expand_grid(
  slope_disp_mean = c(-1, 0, 1),
  disp0_mean = c(0.1, 0.2),
  heritability=c(0, 0.3, 1),
  variance_pheno=c(0,1),
  reproduction=c("\"clonal\"", "\"sexual\""),  ### note how string variable/csts must be entered to fit in later
  ) %>%
  filter(-(heritability != 0 & variance_pheno==0)) %>% ##remove "duplicate" cells
  #(because if V_P=0, then any h^2 will give the same V_A than h^2=0)
  mutate(
    fecundity = 5, ## more or less arbitrary
    K = 500
    ) %>%
  mutate(
    v_f = 2 * sqrt( ## v_f = expected velocity of "equivalent" pulled wave:
      log(fecundity) * (1 - 1 / K) * ## fecundity term at N = 1
        0.5 * invlogit(logit(disp0_mean) + (1 / K) * slope_disp_mean)
    ) ## dispersal term at N = 1
  )
```


Now we need to add all of this (plus information on the variables that stay constant, which output variables will we save...) to the experiment slot of the NL object (we encourage readers to look at the info pane in the .nlogo model file for details):

```{r set-experiment}

nl_object@experiment <- experiment(
  expname = "experiment-h2-2020",
  outpath = here("NetLogo_output"),
  repetition = 1,
  tickmetrics = "true",
  idsetup = "setup",
  idgo = "go",
  stopcond = "not any? turtles",
  runtime = duration,
  evalticks = duration, #test run
  metrics = c("ticks","present_front","new_front"),
  metrics.patches = c("pxcor", "N_predispersal", "N_postdispersal", "N_allele0", "N_allele1"),
  ## we need to add the individual metrics to save here
  constants = list(
    "duration" = duration
  ),
  variables = list( ### needs to be a nested list
    "disp0_mean" = list(values = trt_grid$disp0_mean),
    "slope_disp_mean" = list(values = trt_grid$slope_disp_mean),
    "heritability"=list(values = trt_grid$heritability),
    "variance_pheno_logit_disp0" = list(values = trt_grid$variance_pheno),
    "variance_pheno_disp_slope" = list(values = trt_grid$variance_pheno),
    "fecundity" = list(values = trt_grid$fecundity),
    "K" = list(values = trt_grid$K),
    "reproduction" = list(values = trt_grid$reproduction)
  )
)
```


And we're go for running the simulations! We set a "distinct" simulation design (see `nlrx` help).
(Careful, script is not parallelized, so running <!--the final version--> is expected to take up to a few hours)

```{r running-simulations}

#set.seed(1989) ## we can set a seed here to guarantee the seeds selected below are the same everytime
nl_object@simdesign <- simdesign_distinct(nl = nl_object, nseeds = nreplicates)
nl_object@simdesign@simseeds<-1:nreplicates #or we can rewrite the nseeds here (to guarantee they are all different)
length(unique(nl_object@simdesign@simseeds))==nreplicates #either way, check that all generated seeds are different
exp_h2 <- run_nl_all(nl_object)
```

<!--stop running from here for now, code not yet adapted starting from here-->
# Output

We can now process the result for the main analyses:

```{r processing-output-presaving}

tab <- exp_h2 %>%
  dplyr::select(ticks, slope_disp_mean, disp0_mean, fecundity, K, seedID = `random-seed`, metrics.patches) %>%
  mutate(metrics.patches2 = map(metrics.patches,
    .f = ~ .x %>%
      filter(N_postdispersal > 0) %>% ## we only keep populated patches...
      filter(pxcor == max(pxcor) | pxcor == 0) %>% ## we only keep the edge and the starting patch
      mutate(is.edge = (pxcor == max(pxcor))) %>%
      mutate(
        Hexp = 2 * (N_allele0 / N_postdispersal) * (N_allele1 / N_postdispersal), # expected heterozygosity
        P0 = (N_allele0 / N_postdispersal),
        P1 = (N_allele1 / N_postdispersal)
      )
  )) %>%
  left_join(trt_grid) %>%
  select(-metrics.patches) %>%
  unnest(cols = c(metrics.patches2)) %>%
  mutate(replicateID = paste(K, treatment, seedID)) %>%
  mutate(Location_full = paste(replicateID, pxcor, ticks))
```


```{r saving-output}
write_csv(tab, path = here("NetLogo_output/model-output.csv"))
```

We can now use the `ibm-analysis` scripts to analyse the model outputs.
